---
title:  "Applied Example: Foreign Policy Alignment Among Major Powers"
author: "Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Applied Example: Foreign Policy Alignment Among Major Powers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
HAS_PS   <- requireNamespace("peacesciencer", quietly = TRUE)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = TRUE, warning = TRUE,
  eval      = NOT_CRAN && HAS_PS,
  purl      = FALSE
)
```

## 1 Overview

This vignette walks through a complete `dbn` workflow on real data. We
fit a dynamic model to foreign policy alignment among 12 major powers
(G7 + BRICS) and use impulse response functions (IRFs) to study how
shocks propagate through the network. See `vignette("methodology")` for
the formal framework.

The data here are UNGA voting similarity scores pulled from the
`peacesciencer` package, used purely as a convenient source. The dbn
modelling pipeline is data-source-agnostic: you can build the same
`[n, n, 1, T]` array from any dyad-year frame, for example via
[netify](https://github.com/netify-dev/netify) or
`as_dbn_array_edgelist()`. The DBN model captures both the time-varying
latent structure of the alignment network and the transition dynamics
that govern how it evolves. IRFs then let us ask counterfactual
questions that standard dyadic approaches cannot address, such as what
happens to the broader network if USA-Russia alignment suddenly
improves.

**Requirements.** This vignette needs the `peacesciencer` package and
its external data:

```r
install.packages("peacesciencer")
peacesciencer::download_extdata()
```

When `peacesciencer` is installed the chunks below evaluate during the
vignette build; otherwise they are skipped and rendered as code only.

## 2 Data preparation

We use the `kappavv` measure from `peacesciencer::add_fpsim()`, which
captures pairwise agreement in UNGA voting. It is continuous, symmetric
($\kappa(i,j) = \kappa(j,i)$), and ranges from $-1$ (systematic
disagreement) to $+1$ (perfect agreement). We then build the 4D array
by hand with the standard pattern (loop over rows of the dyad-year
frame); that is the same operation any data source would need.

```{r data-prep}
library(dbn)
library(ggplot2)
library(peacesciencer)

# G7 + BRICS: 12 actors
actor_codes = c(
  USA = 2, Canada = 20, UK = 200, France = 220, Germany = 255,
  Italy = 325, Japan = 740, Brazil = 140, Russia = 365,
  India = 750, China = 710, SouthAfrica = 560
)
actor_labels = names(actor_codes)
n = length(actor_codes)

usa_idx    = which(actor_labels == "USA")
russia_idx = which(actor_labels == "Russia")
china_idx  = which(actor_labels == "China")

# build dyadic panel 2005-2015
dyads = create_dyadyears(subset_years = 2005:2015) |>
  add_fpsim()

dyads_sub = dyads[dyads$ccode1 %in% actor_codes &
                    dyads$ccode2 %in% actor_codes, ]

years = sort(unique(dyads_sub$year))
Tt = length(years)

# build [n, n, 1, T] array: loop over rows of the dyad-year frame.
# this is the canonical pattern for any tidy edgelist source;
# `as_dbn_array_edgelist()` or `netify` would do the same work.
code_to_idx = setNames(seq_along(actor_codes), actor_codes)
Y = array(NA_real_, dim = c(n, n, 1, Tt),
          dimnames = list(actor_labels, actor_labels, "unga",
                          as.character(years)))
for (i in seq_len(nrow(dyads_sub))) {
  row_i = code_to_idx[as.character(dyads_sub$ccode1[i])]
  col_j = code_to_idx[as.character(dyads_sub$ccode2[i])]
  t_idx = which(years == dyads_sub$year[i])
  Y[row_i, col_j, 1, t_idx] = dyads_sub$kappavv[i]
}
for (t in 1:Tt) diag(Y[, , 1, t]) = NA
```

## 3 Fit the dynamic model

`kappavv` is symmetric, so we set `symmetric = TRUE` to constrain
$B_t = A_t$ and halve the parameter count. The sender-receiver
distinction has no meaning for undirected voting similarity.

```{r fit-model}
# vignette settings keep the build short. raise nscan to 20000 and burn
# to 10000 for longer chains. ar1 = TRUE adds a mean-reversion parameter
# to the random walk on A_t so the posterior lag operator stays
# contractive over the horizons we forecast.
fit = dbn(
  data       = Y,
  model      = "dynamic",
  family     = "gaussian",
  symmetric  = TRUE,
  ar1        = TRUE,
  update_rho = TRUE,
  nscan      = 3000,
  burn       = 1500,
  odens      = 5,
  verbose    = FALSE
)
```

### Convergence

Always check convergence before interpreting results. Well-behaved
traces show stationary fluctuation around a stable mean. Under the
symmetric constraint, `tau_A2` and `tau_B2` should be identical.

```{r convergence}
check_convergence(fit)
plot_trace(fit)
```

## 4 Posterior alignment scores

The model estimates a time-varying $\Theta_t$ and a static baseline $M$.
Their sum gives smoothed alignment trajectories for each dyad,
filtering out observation noise so the underlying relational dynamics
come through.

```{r alignment}
Theta_mean = apply(fit$Theta, 1:4, mean)
M_mean = apply(fit$M, 1:3, mean)

alignment = array(NA, dim = c(n, n, 1, length(fit$time_kept)))
for (t_idx in seq_along(fit$time_kept)) {
  alignment[, , 1, t_idx] = Theta_mean[, , 1, t_idx] + M_mean[, , 1]
}

key_dyads = list(
  "USA - UK"       = c("USA", "UK"),
  "USA - China"    = c("USA", "China"),
  "USA - Russia"   = c("USA", "Russia"),
  "China - Russia" = c("China", "Russia"),
  "India - China"  = c("India", "China"),
  "Brazil - USA"   = c("Brazil", "USA")
)

stored_years = years[fit$time_kept]
df_align = do.call(rbind, lapply(names(key_dyads), function(nm) {
  pair = key_dyads[[nm]]
  i = which(actor_labels == pair[1])
  j = which(actor_labels == pair[2])
  data.frame(dyad = nm, year = stored_years,
             alignment = alignment[i, j, 1, ])
}))

ggplot(df_align, aes(x = year, y = alignment, colour = dyad)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Posterior Mean Foreign Policy Alignment",
    subtitle = "G7 + BRICS, UNGA Voting Similarity (2005-2015)",
    x = "Year", y = "Estimated Alignment", colour = "Dyad"
  ) +
  theme_bw() +
  theme(
    panel.border = element_blank(),
    legend.position = "bottom"
  )
```

The model recovers patterns consistent with well-known geopolitical
structures. China-Russia, India-China, and the Western allied dyad
USA-UK sit at positive alignment throughout the panel, while USA-China,
USA-Russia, and Brazil-USA stay negative or near zero, reflecting
systematic disagreement on UNGA votes. Within-dyad trends (USA-UK
drifting upward as the panel matures, for instance) are the kind of
dynamic structure this model exists to surface.

## 5 Impulse response: edge shock

How does a bilateral shock propagate through the broader network? We
simulate a positive shock to USA-Russia alignment and track network
density. Because the data are symmetric, the shock matrix is
symmetrised so it stays consistent with the fit.

```{r irf-edge}
S_edge = build_shock(m = n, type = "unit_edge",
                     i = usa_idx, j = russia_idx, magnitude = 0.5)
S_edge = S_edge + t(S_edge)

irf_density = compute_irf(
  fit,
  shock    = S_edge,
  H        = 5,
  t0       = 3,
  stat_fun = dbn::stat_density,
  n_draws  = 100
)

plot(irf_density, ci_level = 0.90,
     title = "USA-Russia Alignment Shock (+0.5) -> Network Density")
print(irf_density)
```

At horizon 0 the effect is exact: $2 \times 0.5 / 132 \approx 0.0076$.
At later horizons the credible intervals widen, reflecting posterior
uncertainty in $A_t$.

## 6 Impulse response: node shock

What if China shifts alignment toward all partners at once? We apply a
node-out shock and track both network density and China's out-degree
to see whether the effect stays localised or propagates.

```{r irf-node}
S_node = build_shock(m = n, type = "node_out",
                     i = china_idx, magnitude = 0.3)
diag(S_node) = 0
S_node = S_node + t(S_node)

irf_china_dens = compute_irf(
  fit, shock = S_node, H = 5, t0 = 3,
  stat_fun = dbn::stat_density,
  n_draws  = 100
)

irf_china_outdeg = compute_irf(
  fit, shock = S_node, H = 5, t0 = 3,
  stat_fun = function(X) dbn::stat_out_degree(X)[china_idx],
  n_draws  = 100
)

df_irfs = rbind(
  data.frame(irf_china_dens, stat = "Network Density"),
  data.frame(irf_china_outdeg, stat = "China Out-Degree")
)

ggplot(df_irfs, aes(x = horizon, y = mean)) +
  geom_ribbon(aes(ymin = q10, ymax = q90), alpha = 0.2) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ stat, scales = "free_y") +
  labs(
    title = "Node Shock: China Alignment +0.3 (All Partners)",
    x = "Horizon (Years)", y = "IRF (Change From Baseline)"
  ) +
  theme_bw() +
  theme(
    panel.border = element_blank(),
    strip.background = element_rect(fill = "black", color = "black"),
    strip.text.x = element_text(color = "white", hjust = 0)
  )
```

## 7 Comparative IRF: G7 node shocks

Which country's engagement shift produces the largest ripple? We apply
identical outward shocks from each G7 member and compare the resulting
density trajectories. All countries produce the same horizon-0 effect
by construction; differences at later horizons reflect each country's
structural position in the estimated dynamics.

```{r irf-compare}
g7 = c("USA", "Canada", "UK", "France", "Germany", "Italy", "Japan")

irf_list = lapply(setNames(g7, g7), function(country) {
  idx = which(actor_labels == country)
  S = build_shock(m = n, type = "node_out", i = idx, magnitude = 0.3)
  diag(S) = 0
  S = S + t(S)
  compute_irf(fit, shock = S, H = 5, t0 = 3,
              stat_fun = dbn::stat_density, n_draws = 80)
})

df_compare = do.call(rbind, lapply(names(irf_list), function(nm) {
  d = irf_list[[nm]]
  data.frame(horizon = d$horizon, mean = d$mean,
             q10 = d$q10, q90 = d$q90, country = nm)
}))

ggplot(df_compare, aes(x = horizon, y = mean, colour = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Comparative IRF: G7 Node Shocks",
    subtitle = "Same Magnitude; Differences Reflect Structural Position",
    x = "Horizon (Years)", y = "Change in Network Density",
    colour = "Shock Origin"
  ) +
  theme_bw() +
  theme(
    panel.border = element_blank(),
    legend.position = "bottom"
  )
```

Posterior uncertainty (omitted for visual clarity) is typically large
relative to the between-country differences, so read these comparisons
as suggestive rather than definitive without longer chains.

## 8 Clustering analysis

Reciprocity is uninformative under the symmetric constraint
$B_t = A_t$: every $\Theta_t$ is symmetric by construction, so there is
no directional asymmetry to measure and `stat_reciprocity()` returns
`NA` by design. Transitivity (the clustering coefficient)
is the substantive nonlinear summary for symmetric alignment data: it
captures whether a shock between two countries propagates into closed
triangles, the hallmark of bloc formation.

```{r irf-trans}
S_usachn = build_shock(m = n, type = "unit_edge",
                       i = usa_idx, j = china_idx, magnitude = 0.5)
S_usachn = S_usachn + t(S_usachn)

# kappavv is continuous; we threshold each posterior network slice at
# the off-diagonal median of |alignment| so roughly half the dyads count
# as "ties". This keeps the network dense enough that a single
# perturbed edge can close triangles and the IRF stays informative.
align_off_diag = c(alignment[, , 1, ])
align_off_diag = align_off_diag[!is.na(align_off_diag)]
align_median = median(abs(align_off_diag))

irf_trans = compute_irf(
  fit, shock = S_usachn, H = 5, t0 = 3,
  stat_fun = function(X) dbn::stat_transitivity(X, threshold = align_median),
  n_draws = 100
)

plot(irf_trans, ci_level = 0.90,
     title = "USA-China Alignment Thaw (+0.5) -> Network Transitivity")
```

## 9 Practical guidance

### Model settings

For stable credible intervals, use longer chains:

```{r pub-settings, eval = FALSE}
# not evaluated in the vignette build (10000 iters takes several minutes
# on a 12-actor x 11-time-point panel). the demo fit above (nscan = 3000)
# already gives stable point estimates; these longer chains tighten the
# CIs. keep ar1 = TRUE to match the demo fit.
fit_long = dbn(Y, model = "dynamic", family = "gaussian",
               symmetric = TRUE, ar1 = TRUE, update_rho = TRUE,
               nscan = 10000, burn = 5000, odens = 5)
```

### Runtime expectations

| Actors | Time Points | Approx. Time (8 Cores) |
|--------|-------------|------------------------|
| 10-15  | 10-20       | 5-15 minutes           |
| 15-25  | 10-20       | 15-60 minutes          |
| 25-50  | 10-20       | 1-4 hours              |
| 50+    | any         | Use `model = "lowrank"` |

### Interpreting IRFs

Horizon 0 captures the mechanical consequence of the shock. For linear
statistics (density, degree), the horizon-0 effect is exact; for
nonlinear statistics (transitivity, reciprocity on directed data), it
is uncertain even at horizon 0. The decay rate at later horizons is
set by the spectral radius of the Kronecker operator $B_t \otimes A_t$
(under the symmetric constraint, this reduces to $\rho(A_t)^2$). The
credible intervals reflect posterior uncertainty in the model
parameters, not innovation noise. IRFs are counterfactual scenarios
that inject a reduced-form perturbation of the latent state $\Theta_t$
and assume the transition dynamics are invariant to the shock.

### Symmetric networks

Undirected data needs two consistent choices. First, set
`symmetric = TRUE` in `dbn()` to constrain $B_t = A_t$. Second,
symmetrise any shock matrix after construction: `S = S + t(S)`.

### Time-thinned models

When `time_thin > 1`, stored arrays correspond to a subset of the time
series. Map back to original years with `years[fit$time_kept]`.
