---
title:  "Dynamic bilinear network models"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Dynamic bilinear network models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE, comment = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = TRUE, warning = FALSE,
  eval      = NOT_CRAN
)
library(dbn)
library(ggplot2)
```

## 1 The model

The dynamic bilinear model lets the influence matrices $A_t$ and $B_t$
change over time, so it can handle settings where the rules of evolution
themselves evolve:

$$\Theta_t = M + A_t \, (\Theta_{t-1} - M) \, B_t^\top + \varepsilon_t$$

Entry $a_{i,i',t}$ of the sender influence matrix $A_t$ says how
strongly actor $i'$'s past sending behaviour predicts actor $i$'s
current sending at time $t$. $B_t$ does the same on the targeting side.
The baseline mean $M$ absorbs stable dyad-specific tendencies that
persist regardless of the dynamics.

Two more parameters govern how quickly the influence structure
reorganises. Innovation variances $\tau_A^2$ and $\tau_B^2$ control
step-to-step variation in $A_t$ and $B_t$: near zero means a nearly
static structure; larger values allow rapid reorganisation. When
`ar1 = TRUE`, the model adds persistence parameters $\rho_A$ and
$\rho_B$ that measure mean reversion. Values near 1 mean the influence
structure is highly persistent; values near 0 mean rapid reversion to
the prior mean.

## 2 Simulate data

We generate a Gaussian network with time-varying influence matrices.
Setting `ar1 = TRUE` gives the influence matrices AR(1) dynamics, so
each period's influence structure is a noisy version of the previous
period's rather than an uncorrelated random walk. Persistence
$\rho = 0.8$ and innovation variance $\tau^2 = 0.1$ create dynamics
that evolve smoothly and are well-identified at this panel size.

```{r simulate}
sim = simulate_dynamic_dbn(
  n      = 12,
  p      = 1,
  time   = 25,
  sigma2 = 0.3,
  tauA2  = 0.1,
  tauB2  = 0.1,
  ar1    = TRUE,
  rhoA   = 0.8,
  rhoB   = 0.8,
  seed   = 6886
)

Y = sim$Z
dim(Y)
```

## 3 Fit the model

Set `ar1 = TRUE` so the model uses AR(1) dynamics for $A_t$ and $B_t$,
and `update_rho = TRUE` so the persistence parameters are estimated
from the data rather than held at a default.

```{r fit, cache = TRUE}
fit = dbn(
  Y,
  family     = "gaussian",
  model      = "dynamic",
  nscan      = 4000,
  burn       = 2000,
  odens      = 4,
  ar1        = TRUE,
  update_rho = TRUE,
  verbose    = FALSE
)
```

## 4 Convergence diagnostics

The parameters worth watching for the dynamic model are $\sigma^2$
(process noise), $\tau_A^2$ and $\tau_B^2$ (innovation variances), and
$\rho_A$ / $\rho_B$ (persistence). Trace plots should show stationary
fluctuation around a stable mean.

```{r convergence}
check_convergence(fit)
```

```{r trace, fig.height = 5}
plot_trace(fit, pars = c("sigma2", "tau_A2", "tau_B2", "rhoA", "rhoB"))
```

ESS values in the several-hundreds and Geweke z-scores within $\pm 2$
indicate adequate mixing for the scalar parameters. To check
between-chain convergence, run multiple chains and check
rank-normalised split-$\hat R$ (Vehtari et al. 2021):

```{r multichain, eval = FALSE}
mc = dbn_multichain(Y, chains = 4, seeds = c(1, 2, 3, 4),
                    model = "dynamic", family = "gaussian",
                    nscan = 5000, burn = 2500, odens = 5,
                    ar1 = TRUE, update_rho = TRUE, verbose = FALSE)
check_convergence(mc)   # cross-chain Rhat, bulk-ESS, tail-ESS
posterior::as_draws(mc) # for bayesplot / tidybayes downstream
```

## 5 Parameter recovery

### Scalar parameters

```{r params}
param_summary(fit)
```

The model estimates variance parameters that govern the dynamics. The
internal parameterisation may differ in scale from the simulation
inputs, so directly comparing scalar values is not always meaningful.
The cleaner checks are (a) whether the AR(1) persistence parameters
recover the truth and (b) whether the latent network trajectories come
back.

### AR(1) persistence

The simulator set $\rho_A = \rho_B = 0.8$. The posterior should land
near 0.8 with the 95% interval covering it:

```{r rho-recovery}
data.frame(
  param = c("rhoA", "rhoB"),
  truth = c(0.8, 0.8),
  post_mean = c(mean(fit$rhoA), mean(fit$rhoB)),
  q025 = c(quantile(fit$rhoA, 0.025), quantile(fit$rhoB, 0.025)),
  q975 = c(quantile(fit$rhoA, 0.975), quantile(fit$rhoB, 0.975))
)
```

### Latent network recovery

The model's primary output is the latent network state $\Theta_t$. We
compare the posterior mean of $\Theta_t + M$ to the true values across
all dyads and time points. A strong correlation along the diagonal
means the model is picking up both the cross-sectional structure and
the temporal dynamics.

```{r theta-recovery, fig.height = 4.5}
Theta_mean = apply(fit$Theta, 1:4, mean)
M_mean = apply(fit$M, 1:3, mean)

n_act = dim(sim$Theta)[1]
Tt = dim(sim$Theta)[4]

true_vals = numeric(0)
est_vals  = numeric(0)

for (t in seq_len(Tt)) {
  true_mat = sim$Theta[, , 1, t]
  est_mat  = Theta_mean[, , 1, t] + M_mean[, , 1]
  mask = !is.na(true_mat)
  true_vals = c(true_vals, true_mat[mask])
  est_vals  = c(est_vals, est_mat[mask])
}

df_theta = data.frame(truth = true_vals, estimate = est_vals)
r_sq = round(cor(df_theta$truth, df_theta$estimate)^2, 3)

ggplot(df_theta, aes(x = truth, y = estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.1, size = 0.8) +
  labs(
    title    = "Recovery of Latent Network State",
    subtitle = paste0("R-Squared = ", r_sq, " (All Dyads x Time Points)"),
    x = expression("True " * Theta[t]),
    y = expression("Posterior Mean of " * (Theta[t] + M))
  ) +
  theme_bw() +
  theme(panel.border = element_blank())
```

Each point is one off-diagonal dyad at one time point. Some scatter is
expected from posterior shrinkage and estimation uncertainty, but the
bulk of points should cluster along the diagonal.

## 6 Dyad trajectories

`dyad_path()` tracks a specific bilateral relationship over time, with
a ribbon for the posterior credible interval. Wider ribbons mean
greater uncertainty about that dyad's trajectory.

```{r dyad-path, fig.height = 3.5}
dyad_path(fit, i = 1, j = 2)
```

## 7 Forecasting

We forecast 3 steps ahead. `predict()` propagates the operator forward
with the same dynamics the model was fit under. This fit is AR(1), so
the operator mean-reverts: $A_{T+h} = \rho_A A_{T+h-1} + \eta_h^A$,
$\eta_h^A \sim \mathcal{N}(0, \tau_A^2)$, and likewise for $B$ (when
`ar1 = FALSE` this reduces to a pure random walk, $\rho = 1$). Each
posterior draw gives a different operator trajectory on top of a
different innovation realisation. (The ALS fallback path holds $A_T$,
$B_T$ constant and uses bootstrap draws for the operator; see
`?dbn_als`.) Each of the 200 draws below produces a forecast trajectory
that captures all three sources of uncertainty: the operator path, the
parameters, and $\sigma^2$.

```{r forecast}
fc_mean = predict(fit, H = 3, draws = 200, summary = "mean")
dim(fc_mean)
```

The resulting array has class `dbn_forecast` and dimensions
`[actors, actors, relations, horizon]`.

For credible intervals, pass `level = N` (percent) or
`summary = "ci"` with explicit quantiles. The return is a
`dbn_forecast_ci` with `$mean`, `$lower`, and `$upper` arrays:

```{r forecast-ci}
fc_ci = predict(fit, H = 3, draws = 200, level = 90)
summary(fc_ci)
```

Forecasts are most informative over short horizons. At longer horizons
the lag operator acts repeatedly on the latent state and innovation and
parameter uncertainty accumulate, so the predictive bands widen even
though the operator stays contractive (both the AR(1) dynamics and the
default `shrink_rho` prior keep it so). The `lower` / `upper` columns
from `summary(fc_ci)` above show that spread per horizon; check the gap
between them at the last horizon you want to report and stop where it
has grown past what is useful.

## 8 Network statistics over time

`network_summary()` tracks a network-level statistic across time with
posterior credible intervals, giving you a macro-level view of how the
network evolves. For the Gaussian family "density" is the share of
latent ties above zero, so values sit near 0.5 on mean-zero simulated
data rather than near a sparse binary edge density.

```{r network-stats}
ns = network_summary(fit, stat = "density")
head(ns)
```

`edge_prob()` returns the posterior probability that each edge is
positive at a given time point. Values near 1 are reliably positive
edges; values near 0.5 mean the sign is essentially undetermined. The
diagonal is `NA` because self-loops are not modeled.

```{r edge-prob}
ep = edge_prob(fit, rel = 1, time = dim(Y)[4])
round(ep[1:5, 1:5], 2)
```

## 9 Bipartite networks

The package detects bipartite networks automatically when sender and
receiver dimensions differ. A rectangular array sends the model down
the bipartite path: $A$ is estimated as `n_row x n_row` (sender
dynamics) and $B$ as `n_col x n_col` (receiver dynamics).

```{r bipartite, cache = TRUE}
sim_bp = simulate_dynamic_dbn(
  n = 8, n_col = 6, p = 1, time = 12, seed = 6886
)

fit_bp = dbn(
  sim_bp$Z,
  model   = "dynamic",
  family  = "gaussian",
  nscan   = 1000,
  burn    = 500,
  odens   = 2,
  verbose = FALSE
)
summary(fit_bp)
```

## 10 Parallel computation

The dynamic model parallelises row-wise FFBS updates via OpenMP. For
networks with 15 or more actors, multi-threading buys you a meaningful
speedup (typically 2-4x with 4-8 cores).

```{r threads, eval = FALSE}
set_dbn_threads(parallel::detectCores(logical = FALSE))
```

## 11 Scaling guidance

For larger networks the binding constraint is RAM. Use
`estimate_memory()` before fitting, and tune `time_thin` and `odens` to
manage memory.

```{r memory-est}
estimate_memory(
  n_row = 50, n_col = 50, p = 1, Tt = 30,
  nscan = 1000, burn = 500, odens = 2,
  time_thin = 2, family = "gaussian"
)
```

For 100+ actors, expect to need 16-32 GB depending on settings.
`time_thin` and `odens` are the most effective levers for reducing it.

## 12 Next steps

For impulse response analysis that traces how shocks propagate through
the estimated dynamics, see `vignette("impulse_response")`. For known
structural breaks where the influence structure shifts at discrete
time points, see `vignette("piecewise_dbn")`. For large networks
(50+ actors), the low-rank model `dbn(..., model = "lowrank")`
factorises the sender dynamics to keep estimation tractable.
