---
title:  "Regime-switching (HMM) bilinear network models"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Regime-switching (HMM) bilinear network models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = FALSE, warning = FALSE
)
library(dbn)
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
} else {
  message("Install 'ggplot2' to see plots in this vignette.")
}
```

----

**NOTE: This vignette is under construction.** The HMM model variant is
implemented and functional, but the documentation and examples below are
preliminary and subject to change.

----

## 1 Overview

The HMM (Hidden Markov Model) variant of the dynamic bilinear network model
allows for **regime-switching** dynamics. At each time point, the network is
governed by one of *R* discrete regimes, each with its own sender and receiver
effect matrices $A_r$ and $B_r$. Transitions between regimes follow a Markov
chain with an estimated transition matrix $\Pi$.

Use `model = "hmm"` when you expect **abrupt structural breaks** in network
dynamics rather than smooth evolution.

## 2 Simulate a regime-switching network

```{r sim-hmm, cache = TRUE}
sim <- simulate_hmm_dbn(
  n = 12, p = 1, time = 30,
  R = 2,
  sigma2 = 0.3,
  tau_A2 = 0.15, tau_B2 = 0.15,
  transition_prob = 0.85,
  seed = 42
)

dim(sim$Y)
table(sim$S)
```

The true regime sequence `sim$S` shows how often each regime is active.

## 3 Fit the HMM model

```{r fit-hmm, cache = TRUE}
fit_hmm <- dbn(
  sim$Y,
  model   = "hmm",
  family  = "ordinal",
  R       = 2,
  nscan   = 1000,
  burn    = 500,
  odens   = 2,
  verbose = FALSE
)
```

## 4 Model summary

```{r summary-hmm}
summary(fit_hmm)
```

The summary displays the posterior mean transition matrix $\Pi$ and
variance parameters for each regime.

## 5 Regime probabilities

The posterior probability of each regime at each time point:

```{r regime-probs}
probs <- regime_probs(fit_hmm)
head(probs)
```

Visualize with a stacked area plot:

```{r regime-plot, fig.height = 3}
plot_regime_probs(fit_hmm)
```

## 6 Model diagnostics

```{r plot-hmm, fig.height = 8}
plot(fit_hmm)
```

Trace plots for scalar parameters:

```{r trace-hmm, fig.height = 5}
plot_trace(fit_hmm)
```

## 7 Dyad trajectory

Track a specific dyad over time with credible intervals:

```{r dyad-hmm, cache = TRUE, fig.height = 3}
dyad_path(fit_hmm, i = 2, j = 7)
```

## 8 Forecasting

Generate multi-step-ahead forecasts:

```{r forecast-hmm}
pred <- predict(fit_hmm, H = 3, draws = 50, summary = "mean")
dim(pred)
```

The predict method samples future regime states from $\Pi$ and propagates
the bilinear dynamics forward.

## 9 When to use HMM

The HMM model is appropriate when:

- Network dynamics exhibit **discrete structural breaks** (e.g., policy changes,
  crises, regime transitions)
- You want to **classify time points** into distinct behavioral regimes
- The number of regimes *R* is relatively small (2--5)

For smoothly evolving dynamics, consider `model = "dynamic"` instead.
For large networks with many nodes, consider `model = "lowrank"`.
