---
title:  "Regime-switching (HMM) bilinear network models"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Regime-switching (HMM) bilinear network models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = TRUE, warning = FALSE,
  eval      = NOT_CRAN
)
library(dbn)
library(ggplot2)
```

## 1 Overview

The HMM variant of the dynamic bilinear network model handles
regime-switching dynamics. At each time point, the network is governed
by one of $R$ discrete regimes, each with its own influence matrices
$A_r$ and $B_r$. Transitions between regimes follow a Markov chain
with an estimated transition matrix $\Pi$:

$$\Theta_t = M + A_{s_t} \, (\Theta_{t-1} - M) \, B_{s_t}^\top + \varepsilon_t, \quad s_t \sim \text{Markov}(\Pi)$$

Use this when network dynamics have abrupt structural breaks and you
want the model to discover when the breaks happened rather than
specifying them up front. For known break points, the piecewise model
has fewer parameters (see `vignette("piecewise_dbn")`).

Regime identification gets easier with bigger networks, longer time
series, and larger structural differences between regimes. For small
networks (roughly $n \leq 10$) or subtle regime changes, piecewise with
known break points is the more reliable choice.

## 2 Simulate a regime-switching network

We simulate a two-regime network with 15 actors over 50 time periods.
A `transition_prob` of 0.85 means each regime persists for an average
of $1/(1 - 0.85) \approx 6.7$ consecutive periods before switching, so
a 50-period series usually visits both regimes several times. Moderate
`tau_A2` and `tau_B2` values (0.3) keep the regimes visibly distinct
while staying in a numerically stable range; very large innovation
variances (≥ 1.0) here cause the variance posteriors to saturate the
sampler's safety ceiling.

```{r simulate, cache = TRUE}
sim = simulate_hmm_dbn(
  n     = 15,
  p     = 1,
  time  = 50,
  R     = 2,
  sigma2          = 0.3,
  tau_A2          = 0.3,
  tau_B2          = 0.3,
  transition_prob = 0.85,
  seed  = 6886
)

dim(sim$Z)

# true regime sequence and frequencies
sim$S
table(sim$S)
```

The true regime labels show which regime generated each time point.

## 3 Fit the HMM model

Set `model = "hmm"` and specify `R = 2` regimes. We fit on the
continuous latent (`sim$Z`) with the Gaussian family for sharper
recovery at this network size; the ordinal path is also supported but
loses cell-level information through the binning. The sampler jointly
estimates the regime-specific dynamics matrices ($A_r$, $B_r$), the
transition matrix $\Pi$, and the regime labels $s_t$.

```{r fit, cache = TRUE}
fit = dbn(
  sim$Z,
  model   = "hmm",
  family  = "gaussian",
  R       = 2,
  nscan   = 3000,
  burn    = 1500,
  odens   = 3,
  verbose = FALSE
)
```

## 4 Model summary and transition matrix

The summary reports the posterior mean transition matrix $\Pi$ and the
variance parameters. The diagonal of $\Pi$ measures regime persistence:
values near 1 mean the regimes stick; large off-diagonals mean frequent
switching.

```{r summary}
summary(fit)
```

The model's regime labels may not match the simulator's; this is the
standard label-switching issue in mixture models and does not affect
substantive results.

## 5 Regime probabilities

The posterior probability of each regime at each time point gives the
model's classification of the series. Values near 0 or 1 mean confident
classification; near 0.5 means ambiguous. Identification depends on how
different the regimes are (controlled by `tau_A2` and `tau_B2` in the
simulator) and how much data lives at each time point (the number of
actors).

```{r regime-probs, fig.height = 3.5}
plot_regime_probs(fit)
```

`regime_probs()` returns the underlying probability matrix for custom
analysis. Each row sums to 1 across regimes.

```{r regime-table}
probs = regime_probs(fit)
head(probs, 10)
```

### Recovery against the true regime sequence

The model's regime labels may be permuted relative to the simulator's,
so we align them by checking both label orderings and keeping the
assignment with higher agreement.

```{r regime-recovery, fig.height = 3.5}
modal = apply(probs, 1, which.max)
true_S = sim$S

# pick the label permutation that best matches the truth. for R = 2
# the only non-identity permutation is the swap; in general we search
# all R! permutations and keep the highest-agreement assignment.
R = ncol(probs)
all_perms = if (requireNamespace("gtools", quietly = TRUE)) {
  gtools::permutations(R, R)
} else {
  # fallback when gtools is not available: identity and reverse only.
  # exhaustive for R = 2, partial for R >= 3.
  rbind(seq_len(R), rev(seq_len(R)))
}
accs = apply(all_perms, 1, function(p) mean(p[modal] == true_S))
best = all_perms[which.max(accs), ]
modal_aligned = best[modal]
acc = max(accs)

df_states = rbind(
  data.frame(time = seq_along(true_S),       regime = factor(true_S),
             source = "True"),
  data.frame(time = seq_along(modal_aligned), regime = factor(modal_aligned),
             source = "Posterior modal")
)

ggplot(df_states, aes(x = time, y = source, fill = regime)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Regime Sequence Recovery",
    subtitle = paste0("Modal agreement = ", round(acc, 2)),
    x = "Time", y = NULL, fill = "Regime"
  ) +
  theme_bw() +
  theme(panel.border = element_blank(), panel.grid = element_blank())
```

Modal agreement well above the 0.5 chance baseline means the model
recovers the regime structure. Disagreements usually cluster near the
true transition points, where the data themselves are genuinely
ambiguous. At the demo settings here ($n = 15$, $T = 50$,
`nscan = 3000`) recovery is solid; at smaller scales (say
$n \leq 12$, $T \leq 30$, `nscan` in the low hundreds), modal agreement
can sit near the chance baseline even when the regimes are genuinely
distinct, because the sampler does not have enough information per
regime to pin down the labels. For substantive HMM inference, aim for
at least $n \geq 15$, $T \geq 40$, `nscan` $\geq 5000$, and run
multiple chains via `dbn_multichain()`.

## 6 Model diagnostics

The default `plot()` method for HMM fits gives a combined diagnostic
view: regime probabilities, the estimated transition matrix, and MCMC
trace plots for monitoring convergence.

```{r plot-hmm, fig.height = 8}
plot(fit)
```

## 7 Dyad trajectory

`dyad_path()` tracks a specific bilateral relationship over time with
posterior credible intervals, averaging across regime assignments.

```{r dyad, cache = TRUE, fig.height = 3.5}
dyad_path(fit, i = 2, j = 5)
```

## 8 Forecasting

The `predict()` method samples future regime states from $\Pi$ and
propagates the regime-specific bilinear dynamics forward. So forecast
uncertainty captures both the posterior uncertainty in $A_r$, $B_r$ and
the uncertainty about which regime is active in future periods.

```{r forecast}
pred = predict(fit, H = 5, draws = 100, summary = "mean")
dim(pred)
```

## 9 When to use HMM

Reach for the HMM model when network dynamics have discrete structural
breaks (policy changes, crises, regime transitions) and you want
data-driven discovery of when the breaks happened. Keep $R$ small (2-5),
and make sure the network has enough actors and time points to identify
the regime structure. For smoothly evolving dynamics without discrete
breaks, `model = "dynamic"` is a better choice. For known break points,
use `model = "piecewise"`. For large networks (50+ actors),
`model = "lowrank"` scales better.
