---
title:  "Impulse response analysis for network models"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Impulse response analysis for network models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = TRUE, warning = FALSE,
  eval      = NOT_CRAN
)
library(dbn)
library(ggplot2)
```

## 1 Overview

Impulse response functions (IRFs) measure how a shock to one part of
the network propagates through the bilinear dynamics over time. Given
the model $\Theta_t = M + A_t \, (\Theta_{t-1} - M) \, B_t' + \varepsilon_t$,
inject a perturbation $S$ at time $t_0$ and track a network-level
statistic over $H$ horizons:

$$\Delta_0 = S, \quad \Delta_h = A_{t_0+h} \, \Delta_{h-1} \, B_{t_0+h}'$$

At each horizon we compare a network statistic (density, degree, ...)
computed on the perturbed versus baseline network. The shock fires
once; later horizons show how the estimated dynamics carry it forward.

## 2 Fit a well-converged dynamic model

IRFs need a fitted model with time-varying $A$ and $B$ matrices. We use
the Gaussian family for clean interpretation and simulate a 10-actor
network.

```{r fit, cache = TRUE}
sim = simulate_dynamic_dbn(
  n = 10, p = 1, time = 15,
  sigma2 = 0.3, tauA2 = 0.05, tauB2 = 0.05,
  ar1 = TRUE, rhoA = 0.8, rhoB = 0.8,
  seed = 6886
)

fit = dbn(
  sim$Z,
  model      = "dynamic",
  family     = "gaussian",
  ar1        = TRUE,
  update_rho = TRUE,
  nscan      = 3000,
  burn       = 1500,
  odens      = 5,
  verbose    = FALSE
)
```

Always check convergence before computing IRFs; poorly converged
chains produce unreliable credible intervals.

```{r check-conv}
check_convergence(fit)
```

Just as important for IRFs, check the operator's per-period gain
$\rho(A_t)\rho(B_t)$ with `dbn_operator()`. A gain below 1 means shocks
decay; above 1 the operator is locally explosive and a shock propagates
instead, with the credible bands widening at longer horizons. Dynamic
fits apply a light contractivity prior by default (`shrink_rho = 0.9`),
which keeps the gain well inside the stable region, so the IRFs below
decay cleanly. Disable it with `shrink_rho = NULL` and the gain will
climb toward 1 on short panels; lower it to regularise more aggressively.

```{r operator-gain}
op = dbn_operator(fit)
summary(op$stability$gain_mean)
```

## 3 Building shock matrices

`build_shock()` creates structured perturbation matrices. Each type
answers a different counterfactual question.

```{r shocks}
m = 10

# shock a single directed edge (actor 1 -> actor 2)
S_edge = build_shock(m, type = "unit_edge", i = 1, j = 2, magnitude = 2)

# shock all outgoing edges from actor 1
S_node = build_shock(m, type = "node_out", i = 1, magnitude = 1)

# uniform shock to all edges
S_dens = build_shock(m, type = "density", magnitude = 0.5)

# custom block shock (e.g., a group-level perturbation)
S_custom = matrix(0, m, m)
S_custom[1:3, 4:6] = 1.0
```

## 4 Edge shock: effect on network density

How does shocking a single edge affect overall network connectivity?
Add 2 units to the $(1,2)$ entry and track density (mean of off-diagonal
entries) over 5 horizons.

```{r irf-density, cache = TRUE, fig.height = 3.5}
irf_dens = compute_irf(
  fit,
  shock    = S_edge,
  H        = 5,
  t0       = 3,
  stat_fun = dbn::stat_density,
  n_draws  = 100
)

plot(irf_dens, title = "Edge Shock (1->2, +2) -> Network Density")
print(irf_dens)
```

At horizon 0 the effect is exact: adding 2 to one off-diagonal cell in
a 10-actor network changes density by $2 / (10 \times 9) \approx 0.022$.
At later horizons, credible intervals widen as posterior uncertainty in
the transition matrices compounds. The decay rate is set by the
spectral properties of the Kronecker operator $B_t \otimes A_t$: when
$\rho(A_t)\rho(B_t) < 1$ the perturbation decays, near unity it is
sustained, and above unity the response grows.

## 5 Node shock: effect on out-degree

What happens when an actor lifts all outgoing ties at once? The
`node_out` shock sets all entries in row $i$ (including the self-loop
diagonal entry) to the specified magnitude.

```{r irf-degree, cache = TRUE, fig.height = 3.5}
irf_deg = compute_irf(
  fit,
  shock    = S_node,
  H        = 5,
  t0       = 3,
  stat_fun = function(X) stat_out_degree(X)[1],
  n_draws  = 100
)

plot(irf_deg, title = "Node-Out Shock (Actor 1, +1) -> Actor 1 Out-Degree")
```

Actor 1's out-degree jumps by 10 at horizon 0 (the shock sets all 10
entries in row 1, including the diagonal, to magnitude 1). The
posterior mean of the effect decays toward zero at later horizons,
while the credible intervals widen as estimation uncertainty in the
transition matrices accumulates across horizons. That widening is
expected in multi-step propagation, not a sign of trouble.

## 6 Edge shock: effect on reciprocity

Reciprocity is a nonlinear statistic. `stat_reciprocity()` returns the
edgewise matched-edge ratio
$\sum_{i \ne j} X_{ij} X_{ji} / \sum_{i \ne j} X_{ij}^2$, which equals
1 for a fully symmetric network and 0 for a fully antisymmetric one;
this matches `igraph::reciprocity(mode = "default")` and
`sna::grecip(g, measure = "edgewise")`. The dyad-census variant
(Wasserman & Faust 1994) is `sna::grecip(g, measure = "dyadic")`. The
triangle-correlation form is available as `stat_reciprocity_cor()`.
Unlike density, reciprocity depends on the baseline network structure,
so there is posterior uncertainty even at horizon 0: the same shock
produces different reciprocity changes depending on the baseline
network, which varies across posterior draws.

```{r irf-recip, cache = TRUE, fig.height = 3.5}
irf_recip = compute_irf(
  fit,
  shock    = S_edge,
  H        = 5,
  t0       = 3,
  stat_fun = stat_reciprocity,
  n_draws  = 100
)

plot(irf_recip, title = "Edge Shock (1->2, +2) -> Network Reciprocity")
```

The nonzero credible interval at horizon 0 shows the distinction in
action: for linear statistics like density, the horizon-0 effect is a
deterministic function of the shock matrix; for nonlinear statistics
it also interacts with the posterior distribution of the baseline
network.

## 7 Comparing multiple shock-statistic pairs

Combining IRF results in a single figure makes it easier to inspect how
different (shock, statistic) pairs decay. Each panel pairs a specific
shock with a specific statistic; the rows are not directly comparable
in magnitude because the statistics are on different scales, so each
panel uses its own y-axis.

```{r compare-irfs, fig.width = 7, fig.height = 4}
irfs = list(
  "Edge shock -> Density"      = irf_dens,
  "Node shock -> Out-degree"   = irf_deg,
  "Edge shock -> Reciprocity"  = irf_recip
)

df_all = do.call(rbind, lapply(names(irfs), function(nm) {
  d = irfs[[nm]]
  data.frame(
    horizon = d$horizon,
    mean    = d$mean,
    q025    = d$q025,
    q975    = d$q975,
    shock   = nm
  )
}))

ggplot(df_all, aes(x = horizon, y = mean)) +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ shock, scales = "free_y") +
  labs(x = "Horizon", y = "IRF", title = "Impulse Response Comparison") +
  theme_bw() +
  theme(
    panel.border = element_blank(),
    strip.background = element_rect(fill = "black", color = "black"),
    strip.text.x = element_text(color = "white", hjust = 0)
  )
```

The pattern to look for across the panels: a clear initial effect that
decays toward zero, with the `q025` / `q975` ribbon widening but
staying bounded as the horizon grows. That bounded widening is the
expected behaviour from a well-identified fit, and it is what the
panels above show.

## 8 Interpreting IRFs

Horizon 0 captures the immediate, mechanical consequence of the shock.
For linear statistics (density, degree) there is no posterior
uncertainty at horizon 0 because the shock is a fixed constant. For
nonlinear statistics (reciprocity, transitivity), there is uncertainty
even at horizon 0 because the statistic interacts with the baseline
network, which varies across posterior draws.

At horizons 1 and beyond, the estimated dynamics propagate the shock
forward. Uncertainty widens because the transition matrices $A_t$ and
$B_t$ are themselves estimated with posterior noise. The decay rate is
set by the spectral radius of the Kronecker operator $B_t \otimes A_t$,
i.e. the product $\rho(A_t)\rho(B_t)$: values below unity yield decay,
values near unity sustain the perturbation. `dbn_operator(fit)`
returns the per-draw operator norms and flags draws in which the
operator is locally non-contractive; pair it with the IRF when
reporting.

One conceptual point: IRFs are counterfactual scenarios that inject a
perturbation $S$ directly into the latent state at $t_0$ and propagate
it through the estimated operators. The shocks are reduced-form
perturbations of $\Theta_t$, not structurally identified innovations
to $\varepsilon_t$, so the framework sits closer to a generalised
impulse response than to a Cholesky-identified SVAR IRF. The credible
intervals reflect posterior uncertainty in the model parameters, not
sampling variability in $\varepsilon_t$. For a predictive-band
interpretation that integrates over innovation noise, use
`predict(fit, level = ...)` rather than `compute_irf()`.

## 9 Available network statistics

The `stat_fun` argument takes any function that maps a matrix to a
scalar. Built-in statistics include `stat_density()` (mean of
off-diagonal entries), `stat_out_degree()` and `stat_in_degree()` (row
and column sums), `stat_reciprocity()` (edgewise matched-edge ratio
$\sum_{i \ne j} X_{ij} X_{ji} / \sum_{i \ne j} X_{ij}^2$, matching the
`igraph::reciprocity` default), and `stat_transitivity()` (clustering
coefficient). For degree-based statistics, wrap to extract a specific
actor:
`function(X) stat_out_degree(X)[i]`.

## 10 Diagnosing unusual IRF output

When an IRF returns `NA` or zero at horizon 0, `debug_irf()` traces a
single posterior draw end to end for one shocked edge. It prints the
operator dimensions, the baseline and shocked density, and the
per-horizon propagation, which is usually enough to localise the
issue. Specify the shocked edge with `shock_i` / `shock_j`:

```{r debug-irf, eval = FALSE}
debug_irf(fit, draw_idx = 1, shock_i = 1, shock_j = 2)
```

## 11 Longer chains

The examples above use moderate settings to keep vignette build times
reasonable. For substantive analysis, bump both the model chain length
and the number of IRF draws:

```{r pub-settings, eval = FALSE}
fit = dbn(Y, model = "dynamic", family = "gaussian",
          nscan = 10000, burn = 5000, odens = 5)

irf = compute_irf(fit, shock = S, H = 10, t0 = 5,
                  stat_fun = dbn::stat_density, n_draws = 500)
```

## 12 Next steps

For a complete applied workflow with real data and IRFs, see
`vignette("applied_ir")`. For piecewise models that compare pre- and
post-regime influence structures, see `vignette("piecewise_dbn")`.
