---
title:  "Impulse response analysis for network models"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Impulse response analysis for network models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = FALSE, warning = FALSE
)
library(dbn)
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
} else {
  message("Install 'ggplot2' to see plots in this vignette.")
}
```

## 1 Overview

Impulse response functions (IRFs) measure how a **shock** to the network
propagates through the bilinear dynamics over time. Given the model
$\Theta_t = A_t \Theta_{t-1} B_t' + M + \varepsilon_t$, we inject a
perturbation $S$ at time $t_0$ and track a network-level statistic
(density, degree, reciprocity, etc.) over $H$ horizons.

## 2 Fit a dynamic model

IRFs require a fitted model with time-varying $A$ and $B$ matrices. We use
a Gaussian family for cleaner interpretation. The example below uses minimal
settings for fast vignette building; see Section 8 for recommended settings
in practice.

```{r sim-fit, cache = TRUE}
sim = simulate_dynamic_dbn(
  n = 10, p = 1, time = 10,
  sigma2 = 0.3, tauA2 = 0.03, tauB2 = 0.03,
  seed = 123
)

fit = dbn(
  sim$Y,
  model   = "dynamic",
  family  = "gaussian",
  nscan   = 300,
  burn    = 150,
  odens   = 1,
  verbose = FALSE
)
```

## 3 Building shock matrices

The `build_shock()` function creates structured perturbation matrices:

```{r shocks}
m = 10

# Shock a single edge (i -> j)
S_edge = build_shock(m, type = "unit_edge", i = 1, j = 2, magnitude = 2)

# Shock all outgoing edges from actor i
S_node = build_shock(m, type = "node_out", i = 1, magnitude = 1)

# Uniform density shock to all edges
S_dens = build_shock(m, type = "density", magnitude = 0.5)

# Custom shock matrix
S_custom = matrix(0, m, m)
S_custom[1:3, 4:6] = 1.0
```

## 4 Unit edge shock -- network density

Track how a single edge shock affects overall network density:

```{r irf-density, cache = TRUE, fig.height = 3.5}
irf_dens = compute_irf(
  fit,
  shock    = S_edge,
  H        = 6,
  t0       = 2,
  stat_fun = dbn::stat_density,
  n_draws  = 20
)

plot(irf_dens, title = "Edge shock -> Density")
print(irf_dens)
```

## 5 Node shock -- out-degree

Track the effect of activating all outgoing edges of actor 1 on that
actor's out-degree:

```{r irf-degree, cache = TRUE, fig.height = 3.5}
irf_deg = compute_irf(
  fit,
  shock    = S_node,
  H        = 6,
  t0       = 2,
  stat_fun = function(X) stat_out_degree(X)[1],
  n_draws  = 20
)

plot(irf_deg, title = "Node-out shock -> Out-degree of actor 1")
```

## 6 Density shock -- reciprocity

How does a uniform perturbation affect reciprocal ties?

```{r irf-recip, cache = TRUE, fig.height = 3.5}
irf_recip = compute_irf(
  fit,
  shock    = S_dens,
  H        = 6,
  t0       = 2,
  stat_fun = stat_reciprocity,
  n_draws  = 20
)

plot(irf_recip, title = "Density shock -> Reciprocity")
```

## 7 Custom shock -- transitivity

```{r irf-trans, cache = TRUE, fig.height = 3.5}
irf_trans = compute_irf(
  fit,
  shock    = S_custom,
  H        = 6,
  t0       = 2,
  stat_fun = stat_transitivity,
  n_draws  = 20
)

plot(irf_trans, title = "Block shock -> Transitivity")
```

## 8 Comparing multiple shocks

Combine IRF results for a faceted comparison:

```{r compare-irfs, fig.width = 7, fig.height = 5}
irfs = list(
  "Edge -> Density"     = irf_dens,
  "Node -> Out-degree"  = irf_deg,
  "Density -> Recipr."  = irf_recip,
  "Block -> Transit."   = irf_trans
)

df_all = do.call(rbind, lapply(names(irfs), function(nm) {
  d = irfs[[nm]]
  data.frame(
    horizon = d$horizon,
    mean    = d$mean,
    q025    = d$q025,
    q975    = d$q975,
    shock   = nm
  )
}))

ggplot(df_all, aes(x = horizon, y = mean)) +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = 0.2) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  facet_wrap(~shock, scales = "free_y") +
  labs(x = "Horizon", y = "IRF", title = "Impulse Response Comparison") +
  theme_bw() +
  theme(
    panel.border = element_blank(),
    strip.background = element_rect(fill = "black", color = "black"),
    strip.text.x = element_text(color = "white", hjust = 0)
  )
```

**Note on settings:** The examples above use small networks (`n = 10`,
`time = 10`) and short MCMC chains (`nscan = 300`) for fast vignette
building. For publication-quality results, increase these settings:

```{r full-settings, eval = FALSE}
# Recommended settings for substantive analysis
sim = simulate_dynamic_dbn(n = 30, p = 1, time = 50, seed = 123)
fit = dbn(sim$Y, model = "dynamic", family = "gaussian",
           nscan = 5000, burn = 2000, odens = 5, verbose = FALSE)

irf = compute_irf(fit, shock = S_edge, H = 20, t0 = 25,
                    stat_fun = dbn::stat_density, n_draws = 200)
```

## 9 Available network statistics

The `stat_fun` argument accepts any function that maps a matrix to a scalar.
Built-in statistics:

| Function | Description |
|----------|-------------|
| `stat_density(X)` | Mean of off-diagonal entries |
| `stat_out_degree(X)` | Row sums (vector) |
| `stat_in_degree(X)` | Column sums (vector) |
| `stat_reciprocity(X)` | Correlation between $X_{ij}$ and $X_{ji}$ |
| `stat_transitivity(X)` | Clustering coefficient |

For degree-based statistics, wrap to extract a specific actor:
`function(X) stat_out_degree(X)[i]`.
