---
title:  "Low-rank bilinear network models"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Low-rank bilinear network models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = FALSE, warning = FALSE
)
library(dbn)
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
} else {
  message("Install 'ggplot2' to see plots in this vignette.")
}
```

----

**NOTE: This vignette is under construction.** The low-rank model variant is
implemented and functional, but the documentation and examples below are
preliminary and subject to change.

----

## 1 Overview

The low-rank model parameterizes the sender dynamics matrix as
$A_t = U \, \text{diag}(\alpha_t) \, U'$, where $U$ is an orthogonal matrix
on the Stiefel manifold and $\alpha_t$ is a vector of time-varying factor
strengths. This reduces the number of free parameters from $m^2$ to
$m \times r + r$, making it **scalable to larger networks** (50+ nodes).

Use `model = "lowrank"` when you have many actors and want an interpretable
factor decomposition of sender dynamics.

## 2 Simulate a low-rank network

We start by simulating a 20-node network over 25 time periods with `r` set to 3. This gives us ground truth to check recovery of the factor structure.

```{r sim-lr, cache = TRUE}
sim = simulate_lowrank_dbn(
  n = 20, p = 1, time = 25,
  r = 3,
  sigma2 = 0.3,
  tau_alpha2 = 0.05,
  tauB2 = 0.03,
  seed = 42
)

dim(sim$Y)
```

## 3 Memory estimation

Low-rank models store the Stiefel matrix $U$ and factor paths at each MCMC iteration, so memory usage can grow quickly with network size. Before fitting, you can estimate RAM requirements:

```{r memory-est}
estimate_memory(
  n_row = 20, n_col = 20, p = 1, Tt = 25,
  nscan = 1000, burn = 500, odens = 2,
  family = "ordinal"
)
```

## 4 Fit the low-rank model

The key argument is `r`, which sets the rank of the factor decomposition. Setting `r` much smaller than `n` is what makes this model tractable for larger networks.

```{r fit-lr, cache = TRUE}
fit_lr = dbn(
  sim$Y,
  model   = "lowrank",
  family  = "ordinal",
  r       = 3,
  nscan   = 1000,
  burn    = 500,
  odens   = 2,
  verbose = FALSE
)
```

## 5 Model summary

The summary shows posterior means for the variance parameters and the estimated rank structure.

```{r summary-lr}
summary(fit_lr)
```

## 6 Diagnostics

The low-rank plot shows MCMC traces, factor trajectories ($\alpha_k$),
and the posterior mean of the node loading matrix $U$:

```{r plot-lr, fig.height = 10}
plot(fit_lr)
```

Scalar parameter traces:

```{r trace-lr, fig.height = 5}
plot_trace(fit_lr)
```

## 7 Factor trajectories

The factor trajectories $\alpha_t$ capture how each latent dimension's influence on sender dynamics evolves over time. A flat trajectory means that factor's contribution is stable; a trending or volatile trajectory suggests structural change along that dimension. Extract factor paths in tidy format for custom plotting:

```{r tidy-lr}
alpha_df = tidy_dbn_lowrank(fit_lr)
head(alpha_df)
```

```{r factor-plot, fig.height = 3}
ggplot(alpha_df, aes(x = time, y = mean)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  geom_line() +
  facet_wrap(~factor, scales = "free_y", labeller = label_bquote(alpha[.(factor)])) +
  labs(x = "Time", y = expression(alpha)) +
  theme_bw() +
  theme(
    panel.border = element_blank(),
    strip.background = element_rect(fill = "black", color = "black"),
    strip.text.x = element_text(color = "white", hjust = 0)
  )
```

## 8 Network summaries

You can compute summary statistics across the posterior distribution of fitted networks. `network_summary()` returns time-varying means or densities, and `edge_prob()` gives the posterior probability that each edge exceeds a threshold.

```{r network-summary}
network_summary(fit_lr, stat = "mean")
network_summary(fit_lr, stat = "density")
edge_prob(fit_lr, threshold = 0)
```

## 9 Dyad trajectory

Track a specific sender-receiver pair over time with posterior credible intervals.

```{r dyad-lr, cache = TRUE, fig.height = 3}
dyad_path(fit_lr, i = 2, j = 7)
```

## 10 Forecasting

`predict()` propagates the low-rank dynamics forward by `H` steps, sampling from the posterior to produce forecast distributions.

```{r forecast-lr}
pred = predict(fit_lr, H = 3, draws = 50, summary = "mean")
dim(pred)
```

## 11 When to use low-rank

The low-rank model is appropriate when:

- The network has **many actors** (50+) making the full dynamic model
  computationally expensive
- You want an **interpretable factor structure** where $U$ reveals actor
  groupings and $\alpha_t$ captures time-varying factor strengths
- The rank $r$ is much smaller than the number of actors

For smaller networks, `model = "dynamic"` provides more flexible dynamics.
For regime-switching behavior, consider `model = "hmm"`.
