---
title:  "Piecewise-static models for structural change"
author: "Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Piecewise-static models for structural change}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = TRUE, warning = FALSE,
  eval      = NOT_CRAN
)
library(dbn)
library(ggplot2)
```

## 1 Overview

> **Terminology note.** Throughout this vignette "block" means a *time
> interval* (a contiguous run of periods sharing the same dynamics),
> not an actor block in the stochastic block model sense. The piecewise
> model partitions the time axis, not the node set.

Many political processes have structural breaks: discrete moments when
the rules of network dynamics fundamentally change. The 2008 financial
crisis reshuffled economic dependencies. The Arab Spring reorganised
regional alliance patterns. A leadership transition can redirect a
country's foreign policy overnight.

The piecewise-static model handles this case directly: you say when the
breaks happened and the model estimates how the influence structure
differs across regimes. It fits block-constant matrices $A_k$ and $B_k$
for regimes $k = 1, \ldots, K$:

$$\Theta_t = M + A_{k(t)} \, (\Theta_{t-1} - M) \, B_{k(t)}^\top + \varepsilon_t$$

Piecewise sits between the static and dynamic extremes. It gives up the
time-point-level precision of the dynamic model in exchange for
interpretability (each regime has a single, directly comparable
influence matrix) and computational efficiency ($K$ matrices instead of
$T-1$).

| Consideration | Piecewise | Dynamic |
|:---|:---:|:---:|
| Breaks are discrete and known | X | |
| Influence evolves smoothly | | X |
| Comparing pre/post periods | X | |
| Want interpretable regime summaries | X | |
| Memory constraints (large networks) | X | |
| Need time-point-specific estimates | | X |

## 2 Simulate data with a structural break

We simulate a network where the influence structure changes at a known
break point, then check whether the model recovers the difference. The
`blocks` argument specifies the ending time index of each block: block 1
covers $t = 1$ to 15, block 2 covers $t = 16$ to 30.

```{r simulate, cache = TRUE}
sim = simulate_piecewise_dbn(
  n      = 12,
  time   = 30,
  blocks = c(15, 30),
  p      = 1,
  sigma2 = 0.5,
  tau2   = 0.3,
  seed   = 6886
)

dim(sim$Y)

# block structure
sim$block_info$K
sim$block_info$boundaries
sim$block_info$lengths

# average element-wise difference between regime A matrices
round(mean(abs(sim$true_A[[1]] - sim$true_A[[2]])), 3)
```

The element-wise difference between the two true $A$ matrices confirms
the data-generating process really does produce different regimes.

## 3 Fit the piecewise model

Pass the known break points via `blocks`. The model estimates separate
$A_k$ and $B_k$ per regime while sharing the baseline mean $M$.

```{r fit, cache = TRUE}
fit = dbn(
  sim$Y_continuous,
  model   = "piecewise",
  family  = "gaussian",
  blocks  = c(15, 30),
  nscan   = 2000,
  burn    = 1000,
  odens   = 2,
  verbose = FALSE
)

summary(fit)
```

## 4 Convergence diagnostics

```{r convergence}
check_convergence(fit)
```

```{r trace, fig.height = 5}
plot_trace(fit, pars = c("s2", "t2", "g2"))
```

## 5 Regime comparison

`compare_blocks()` quantifies the difference between regimes with
posterior uncertainty. It returns the posterior mean of the Frobenius
norm $\|A_k - A_{k+1}\|$, a 95% credible interval, and the posterior
probability that the difference exceeds a substantively meaningful
threshold (default 0.1). A high probability is strong evidence the
influence structure genuinely shifted, not just sampling noise.

```{r compare-blocks}
regime_diffs = compare_blocks(fit)
```

## 6 Extracting regime-specific influence

Each regime has its own $A$ and $B$ matrices stored in the fit object.

```{r regime-matrices}
A1_est = fit$A_blocks[[1]]
A2_est = fit$A_blocks[[2]]
```

`compare_blocks()` is the right tool for assessing whether the
influence structure genuinely changed between regimes. To compare
individual $A$ elements entry by entry, use a larger network
($n \geq 20$).

## 7 Recovering the per-regime operators

The substantive check is whether each regime's $A$ matrix comes back.
We plot estimated vs. true entries for each block.

```{r influence-change, fig.height = 4.5}
df_recover = rbind(
  data.frame(truth = as.vector(sim$true_A[[1]]),
             estimate = as.vector(A1_est), regime = "Regime 1"),
  data.frame(truth = as.vector(sim$true_A[[2]]),
             estimate = as.vector(A2_est), regime = "Regime 2")
)
r_sq <- by(df_recover, df_recover$regime, function(d) {
  round(cor(d$truth, d$estimate)^2, 2)
})

ggplot(df_recover, aes(x = truth, y = estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_point(alpha = 0.5, size = 0.9) +
  facet_wrap(~ regime) +
  labs(
    title    = "Per-regime A recovery",
    subtitle = sprintf("R^2: %s = %s, %s = %s",
                       names(r_sq)[1], r_sq[[1]], names(r_sq)[2], r_sq[[2]]),
    x = "True A entry", y = "Posterior mean A entry"
  ) +
  theme_bw() +
  theme(panel.border = element_blank(),
        strip.background = element_rect(fill = "black", color = "black"),
        strip.text = element_text(color = "white", hjust = 0))
```

Points clustering along the 45-degree line means the model recovers
each regime's operator entry by entry. The block-norm contrast
shown in section 5 (`compare_blocks()`) handles the regime *difference*
with proper posterior uncertainty.

## 7b Impulse response on a piecewise fit

`compute_irf()` works on piecewise fits: under the hood it tiles each
block's `A_k` / `B_k` across the time indices that belong to the block,
producing a per-time operator cube that the standard IRF machinery
consumes. Because the operator differs between regimes, the same shock
applied early (regime 1) versus late (regime 2) produces different
propagation patterns. That contrast is the whole payoff of using a
piecewise fit.

```{r piecewise-irf, fig.height = 3.5}
n = nrow(A1_est)
shock = build_shock(m = n, type = "node_out", i = 1, magnitude = 1)

irf_r1 = compute_irf(fit, shock = shock, H = 4, t0 = 5,
                     n_draws = 100, stat_fun = dbn::stat_density,
                     seed = 6886)
irf_r2 = compute_irf(fit, shock = shock, H = 4, t0 = 22,
                     n_draws = 100, stat_fun = dbn::stat_density,
                     seed = 6886)

df_irf = rbind(
  data.frame(horizon = irf_r1$horizon, mean = irf_r1$mean,
             lo = irf_r1$q025, hi = irf_r1$q975, regime = "Regime 1 (t0=5)"),
  data.frame(horizon = irf_r2$horizon, mean = irf_r2$mean,
             lo = irf_r2$q025, hi = irf_r2$q975, regime = "Regime 2 (t0=22)")
)

ggplot(df_irf, aes(x = horizon, y = mean, colour = regime, fill = regime)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Node-Out Shock to Actor 1 -> Network Density",
    x = "Horizon", y = "Change in Density", colour = "", fill = ""
  ) +
  theme_bw() +
  theme(
    panel.border    = element_blank(),
    legend.position = "bottom"
  )
```

For per-draw diagnostics on a single posterior sample, `debug_irf()`
prints the operator dimensions, the per-draw shock, and the horizon-0
statistic. Reach for it when an unusual shock or fit configuration
returns NA.

## 8 Speed and memory advantages

The piecewise model estimates $K$ influence matrices instead of $T-1$
and pools data within each regime, so it is both faster and lighter on
memory than the dynamic model, especially on long time series.

As a rule of thumb on this data shape, piecewise comes in 2-4x faster
than dynamic at the same `nscan`, with the gap widening as $T$ grows.
The block-Gibbs updater pools all in-block observations into one
$A_k$ / $B_k$ update, so the per-iteration work scales with $K$ rather
than $T-1$.

On large networks, pass `store_theta = FALSE` to skip the full $\Theta$
posterior (which scales as $n^2 \times T \times \text{draws}$). You
keep $A$, $B$, $M$, the variance posteriors, convergence diagnostics,
and `compare_blocks()`; only the per-cell $\Theta$ uncertainty goes
away.

```{r memory-compare}
# same data and family as the main fit above, just store_theta = FALSE
fit_lean = dbn(sim$Y_continuous, model = "piecewise", family = "gaussian",
               blocks = c(15, 30), nscan = 2000, burn = 1000, odens = 2,
               verbose = FALSE, store_theta = FALSE)

data.frame(
  storage = c("With Theta (fit above)", "Without Theta (fit_lean)"),
  mb      = round(c(object.size(fit), object.size(fit_lean)) / 1e6, 2)
)
```

## 9 Specifying blocks

The `blocks` argument is flexible. A single integer creates that many
equal-sized blocks. A vector of integers gives the ending time index of
each block. Named vectors help readability in applied work.

```{r block-specs, eval = FALSE}
# integer: equal-sized regimes
fit = dbn(Y, model = "piecewise", blocks = 4)

# vector of endpoints
fit = dbn(Y, model = "piecewise", blocks = c(10, 25, 40))

# named vector for interpretability
fit = dbn(Y, model = "piecewise",
          blocks = c(pre_war = 15, war = 30, post_war = 45))
```

## 10 Other outcome families

The piecewise model also supports `family = "ordinal"` (for integer
ratings or Likert items, via the rank likelihood) and
`family = "binary"` (probit-augmented). The call pattern is identical;
just change the family and pass the right input scale.

## 11 Applied example: UNGA voting and the 2008 financial crisis

This example examines UN General Assembly voting alignment using the
financial crisis as a structural break, following Salau & Minhas (under
review). It needs the `peacesciencer` package and longer chains than
the vignette build allows, so we show the code and a representative
output rather than evaluating inline. The dbn modelling pipeline is
data-source-agnostic; the `peacesciencer` calls below are just a
convenient way to assemble the dyad-year panel. You could equally
well build `Y_unga` via [netify](https://github.com/netify-dev/netify)
or `as_dbn_array_edgelist()`.

```r
library(peacesciencer)

# build dyadic panel 1995-2015
dyads = create_dyadyears(subset_years = 1995:2015) |>
  add_fpsim()

# select ~100 countries with complete data
years = 1995:2015
complete_countries = Reduce(intersect, lapply(years, function(y) {
  unique(c(dyads$ccode1[dyads$year == y], dyads$ccode2[dyads$year == y]))
}))

country_coverage = table(c(
  dyads$ccode1[dyads$ccode1 %in% complete_countries],
  dyads$ccode2[dyads$ccode2 %in% complete_countries]
))
top_countries = as.numeric(
  names(sort(country_coverage, decreasing = TRUE))[1:100]
)

# build [n, n, 1, T] array of kappavv similarity scores
dyads_sub = dyads[dyads$ccode1 %in% top_countries &
                    dyads$ccode2 %in% top_countries, ]
actor_codes = sort(unique(c(dyads_sub$ccode1, dyads_sub$ccode2)))
n_actors = length(actor_codes)
code_to_idx = setNames(seq_along(actor_codes), actor_codes)

Y_unga = array(NA_real_, dim = c(n_actors, n_actors, 1, length(years)))
for (i in seq_len(nrow(dyads_sub))) {
  row_i = code_to_idx[as.character(dyads_sub$ccode1[i])]
  col_j = code_to_idx[as.character(dyads_sub$ccode2[i])]
  t_idx = which(years == dyads_sub$year[i])
  Y_unga[row_i, col_j, 1, t_idx] = dyads_sub$kappavv[i]
}
for (t in seq_along(years)) diag(Y_unga[, , 1, t]) = NA

# fit piecewise model with a known break (e.g. a financial crisis year)
t_break = which(years == 2008)
fit_crisis = dbn(
  Y_unga,
  model  = "piecewise",
  family = "gaussian",
  blocks = c(t_break, length(years)),
  nscan  = 5000,
  burn   = 2000,
  odens  = 5,
  store_theta = FALSE,
  verbose = TRUE
)

# did the influence structure change?
compare_blocks(fit_crisis)
#> Block Comparison Results (A)
#> block_1 vs block_2: ||ΔA|| = 2.34 [1.89, 2.91]
#>   P(||ΔA|| > 0.1) = 1

# which actors' influence changed most?
A_pre  = fit_crisis$A_blocks[[1]]
A_post = fit_crisis$A_blocks[[2]]
influence_change = rowSums(abs(A_post)) - rowSums(abs(A_pre))
```

A high probability that the Frobenius norm exceeds the threshold
indicates the crisis genuinely altered alignment dynamics. In typical
applications, emerging market countries from Latin America, Asia, and
Africa show increased influence on alignment patterns post-2008, while
G7 nations show flat or declining dynamic influence. Countries that
actively built coalitions (Brazil, Turkey, South Africa) tend to come
out as gainers. Substantive conclusions still need longer MCMC runs,
robustness checks with alternative break points, and domain expertise.

## 12 Scaling to large networks

| Actors | Theta Storage | Expected Runtime | Memory |
|:---:|:---:|:---:|:---:|
| 50 | OK | 1-5 min | < 1 GB |
| 100 | Marginal | 5-20 min | 2-8 GB |
| 200 | Disable | 30-90 min | 1-4 GB |
| 500+ | Disable | Hours | 5-20 GB |

For 200+ actors, always pass `store_theta = FALSE`. You keep $A$, $B$,
$M$, the variance posteriors, convergence diagnostics, and
`compare_blocks()` support; you lose the full $\Theta$ posterior
uncertainty.

```{r large-settings, eval = FALSE}
fit_large = dbn(
  Y_large,
  model       = "piecewise",
  family      = "gaussian",
  blocks      = c(25, 50),
  store_theta = FALSE,
  nscan       = 5000,
  burn        = 2000,
  odens       = 5,
  verbose     = TRUE
)
```

## 13 Practical guidance

### Choosing break points

Good candidates: known events (war onset, treaty signing), policy
interventions (sanctions, trade agreements), and external shocks
(financial crises, pandemics). If you are unsure where the break is,
fit multiple models with different break specifications and compare
with `compare_dbn()`.

### How many regimes?

| Time Points | Max Regimes |
|:---:|:---:|
| 20-30 | 2-3 |
| 30-50 | 3-4 |
| 50-100 | 4-6 |

### Relationship to other models

Piecewise with $K = 1$ collapses to the static model. Compared with the
dynamic model, piecewise gives up time-point precision for
interpretability and speed. Compared with HMM, piecewise asks the
analyst for break points rather than discovering them from the data.

## 14 Next steps

For smoothly evolving dynamics, see `vignette("dynamic_dbn")`. For
data-driven regime discovery, see `vignette("hmm_dbn")`. For impulse
response analysis, see `vignette("impulse_response")`. For a full
applied workflow with IRFs, see `vignette("applied_ir")`.
