---
title:  "Getting started: static bilinear network model"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started: static bilinear network model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = FALSE, warning = FALSE
)
library(dbn)
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
} else {
  message("Install 'ggplot2' to see plots in this vignette.")
}
```

## 1 Simulate a small static network

The static model estimates time-invariant sender and receiver effects.
We start by simulating ordinal relational data.

```{r simulate-data}
set.seed(1)
sim <- simulate_static_dbn(n = 15,            # network size: 15 actors
                          p   =  1,            # single relation type
                          time  =  5)          # 5 replicate time periods
Y <- sim$Y
dim(Y)   # [n_row, n_col, p, time]
```

## 2 Fit the model

```{r fit-model, cache = TRUE}
fit_static <- dbn(sim$Y,
                  model = 'static',
                  family = 'ordinal',   # rank-likelihood for ordinal data
                  nscan = 600,          # MCMC iterations after burn-in
                  burn = 200,           # burn-in period
                  odens = 1             # save every iteration
                  )
```

## 3 Convergence diagnostics

Always check convergence before interpreting results.

```{r convergence}
check_convergence(fit_static)
plot_trace(fit_static, pars = c("s2", "t2", "g2"))
```

## 4 Model summary and parameter inspection

```{r summaries}
summary(fit_static)
param_summary(fit_static)
```

## 5 Latent mean structure

The baseline mean M captures the average relational tendency for each dyad.
Extract and summarize M across posterior draws:

```{r latent-summary}
M_summary <- latent_summary(fit_static, fun = mean)
head(M_summary)
```

## 6 Gaussian and binary families

The static model supports all three outcome families.

**Gaussian data** (continuous outcomes):

```{r gaussian-static, cache = TRUE}
# Simulate continuous data
set.seed(10)
n <- 10; p <- 1; time <- 5
Z_gauss <- array(rnorm(n * n * p * time), dim = c(n, n, p, time))
for (t in 1:time) diag(Z_gauss[,,1,t]) <- NA

fit_gauss <- dbn(Z_gauss, model = "static", family = "gaussian",
                 nscan = 300, burn = 150, verbose = FALSE)
summary(fit_gauss)
```

**Binary data** (0/1 outcomes via probit link):

```{r binary-static, cache = TRUE}
set.seed(20)
Y_bin <- array(rbinom(n * n * p * time, 1, 0.3), dim = c(n, n, p, time))
for (t in 1:time) diag(Y_bin[,,1,t]) <- NA

fit_bin <- dbn(Y_bin, model = "static", family = "binary",
               nscan = 300, burn = 150, verbose = FALSE)
summary(fit_bin)
```

## 7 Wrap-up

* We simulated ordinal network data,
* fit a static bilinear model with MCMC,
* checked convergence,
* inspected parameter posteriors and latent means,
* showed how to switch between ordinal, gaussian, and binary families.
