---
title:  "Getting started: static bilinear network model"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started: static bilinear network model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = TRUE, warning = FALSE,
  eval      = NOT_CRAN
)
library(dbn)
library(ggplot2)
```

## 1 Overview

Dynamic bilinear network (DBN) models estimate how past network
interactions predict future ones. The package takes a time series of
networks stored as a 4D array with dimensions
`[senders, receivers, relation_types, time_points]`, with diagonal
entries (self-ties) set to `NA`.

The static model is the simplest variant. It estimates time-invariant
sender and receiver influence matrices $A$ and $B$ that govern how the
latent network state evolves:

$$\Theta_t = M + A \, (\Theta_{t-1} - M) \, B^\top + \varepsilon_t$$

$A$ captures sender influence: entry $a_{i,i'}$ measures how strongly
actor $i'$'s past sending behaviour predicts actor $i$'s current
sending. $B$ does the same on the receiving side. $M$ absorbs persistent
dyad-specific tendencies like stable rivalries or alliances. Because
$A$ and $B$ are constant across time, the rules of evolution do not
change; only the network state $\Theta_t$ itself moves.

## 2 Simulate data

We simulate a small network so we have known ground truth to compare
against. The simulator returns several pieces; the two we need are the
observed ordinal array (`sim$Y`, integer ratings 1-5) and the underlying
continuous latent (`sim$Z`). We use `sim$Z` here because the Gaussian
family expects continuous input; section 8 below uses `sim$Y` for the
ordinal family.

```{r simulate}
sim = simulate_static_dbn(
  n      = 12,   # 12 actors
  p      = 1,    # 1 relation type
  time   = 15,   # 15 time periods
  sigma2 = 0.5,  # process (state) noise variance
  tau2   = 0.1,  # prior variance for A and B
  seed   = 6886
)

Y = sim$Z
dim(Y)
```

The output confirms the expected `[12, 12, 1, 15]` array structure.  The
simulation also returns the true $A$, $B$, $M$, and $\Theta$ matrices,
which we use below to verify that the model recovers the data-generating
process.

## 3 Fit the model

Pass the data array to `dbn()` and specify the model type and outcome
family. Three MCMC settings control sampling: `nscan` is the number of
posterior samples to draw after burn-in, `burn` is the warm-up to
discard, and `odens` thins the chain by saving every $k$-th sample to
reduce autocorrelation.

```{r fit, cache = TRUE}
fit = dbn(
  Y,
  model   = "static",
  family  = "gaussian",
  nscan   = 2000,
  burn    = 1000,
  odens   = 2,
  verbose = FALSE
)
```

## 4 Convergence diagnostics

MCMC samples from the posterior are only useful once the chain has
converged: it should have found the right region of parameter space and
be exploring it stably. Well-behaved trace plots show stationary
fluctuation around a stable mean. Drift, sticking, or slow oscillation
all suggest the chain needs to run longer.

```{r convergence, fig.height = 5}
check_convergence(fit)
plot_trace(fit)
```

The diagnostics report effective sample size (ESS, via `coda`) and
Geweke statistics. A few hundred ESS is adequate; in well-mixing chains
ESS approaches the number of stored draws. The three parameters shown
are $\sigma^2$ (process noise variance), $\tau^2$ (the joint prior
variance shared by the entries of $A$ and $B$), and $g^2$ (the prior
variance on elements of $M$). If any trace shows trending or poor
mixing, bump both `burn` and `nscan` and rerun before going further.

If you live in the `posterior` / `tidybayes` / `bayesplot`
ecosystem, the fit object supports `as_draws()` directly, giving the
rank-normalised split-$\hat R$, bulk-ESS, and tail-ESS from Vehtari et
al. 2021:

```{r posterior, eval = NOT_CRAN && requireNamespace("posterior", quietly = TRUE)}
posterior::summarise_draws(posterior::as_draws(fit))
```

## 5 Parameter recovery

The point of a simulation study is to check whether the model recovers
the known truth. Scalar variance parameters can differ in scale between
the simulator and the model's internal parameterisation, so the most
informative check is whether the latent baseline structure $M$ comes
back.

### Scalar parameters

```{r param-recovery}
param_summary(fit)
```

### Baseline mean M

The posterior mean of $M$ should correlate with the true $M$ from the
simulation. We exclude diagonal entries (self-ties are `NA` by
construction) and compare all off-diagonal elements.

```{r m-recovery, fig.height = 4.5}
M_est = latent_summary(fit, fun = mean)

# exclude diagonal entries from both truth and estimate
n_act = 12
off_diag = which(M_est$i != M_est$j)
est_M_vec = M_est$value[off_diag]

true_M_mat = sim$M[, , 1]
diag(true_M_mat) = NA
true_M_vec = as.vector(true_M_mat)
true_M_vec = true_M_vec[!is.na(true_M_vec)]

df_M = data.frame(truth = true_M_vec, estimate = est_M_vec)
r_sq = round(cor(df_M$truth, df_M$estimate)^2, 3)

ggplot(df_M, aes(x = truth, y = estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.5) +
  labs(
    title = "Recovery of Baseline Mean M",
    subtitle = paste0("R-Squared = ", r_sq),
    x = "True M", y = "Posterior Mean M"
  ) +
  theme_bw() +
  theme(panel.border = element_blank())
```

Points clustering along the 45-degree line mean the model is recovering
the true baseline relational tendencies. An $R^2$ above 0.8 says the
posterior mean of $M$ is picking up most of the cross-sectional
structure in the data-generating process.

## 6 Model summary

`summary()` reports posterior means and credible intervals for the
scalar parameters, along with data dimensions, MCMC settings, and the
outcome family.

```{r summary}
summary(fit)
```

## 7 Posterior predictive check

A posterior predictive check asks whether data simulated from the fitted
model look like the observed data. We generate replicated datasets from
the posterior, then compare their empirical CDF to the observed data.

```{r ppc, fig.height = 4}
ppd = posterior_predict_dbn(fit, draws = 30, seed = 6886)
plot_ppc_ecdf(fit, ppd, Y_obs = Y, rel = 1)
```

Replicated ECDFs that track the observed curve closely mean the model
is picking up the data's distributional features. Systematic departures
across the bulk of the support point to mis-specification rather than
sampling noise.

## 8 Other outcome families

The static model supports two more families beyond Gaussian. The
ordinal family uses a rank likelihood that respects the ordering of
categorical data (event counts, Likert items) without assuming a
specific scale. The binary family uses a probit link for 0/1 tie data.
Both follow the same call pattern; just switch `family` and pass the
right input scale:

```r
# ordinal: integer-coded ranks
fit_ord = dbn(Y_int, model = "static", family = "ordinal",
              nscan = 2000, burn = 1000, odens = 2)

# binary: 0/1 ties, with NA on the diagonal
fit_bin = dbn(Y_bin, model = "static", family = "binary",
              nscan = 2000, burn = 1000, odens = 2)
```

## 9 Next steps

The static model assumes the influence structure is constant. If those
rules might change, the dynamic model lets $A_t$ and $B_t$ evolve over
time (see `vignette("dynamic_dbn")`). For known structural breaks, the
piecewise model estimates separate influence matrices per regime
(`vignette("piecewise_dbn")`). For the full mathematical framework, see
`vignette("methodology")`.
