---
title:  "Getting started: static bilinear network model"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started: static bilinear network model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = FALSE, warning = FALSE
)
library(dbn)
library(ggplot2)      # needed for visualization functions
```

## 1  Simulate a small static network

```{r simulate-data}
set.seed(1)
sim <- simulate_static_dbn(n = 15,            # network size: 15 actors
                          p   =  1,            # single relation type
                          time  =  1)            # static model = one time period
Y <- sim$Y
dim(Y)
```

## 2  Fit a *very* short MCMC chain

```{r fit-model, cache = TRUE}
fit_static <- dbn(sim$Y,
                  model = 'static',
                  nscan = 600,    # mcmc iterations after burn-in
                  burn = 200,     # initial burn-in iterations
                  odens = 1      # save every iteration (no thinning)
                  )
```

## 3  Step #1 - Assess convergence

```{r convergence}
check_convergence(fit_static)
plot_trace(fit_static, pars = c("s2", "t2", "g2"))
```

## 4  Inspect posterior summaries

```{r summaries}
summary(fit_static)                 # model summary via s3 method
plot(fit_static)                    # default plots via s3 dispatch
```

## 5  Posterior-predictive checks

Generate 100 replicated networks and overlay the ECDFs:

```{r ppc-ecdf, cache = TRUE}
ppd <- posterior_predict_dbn(fit_static, ndraws = 100, seed = 42)
plot_ppc_ecdf(fit_static, ppd, ndraws_plot = 30, alpha = 0.2, Y_obs = Y)
```

Density version (identical behaviour when data are continuous):

```{r ppc-density}
plot_ppc_density(fit_static, ppd, Y_obs = Y)
```

## 6  Drawing *new* predictions

```{r simulate-new, eval = FALSE}
# generate 500 posterior draws - returns array [m × m × p × n × 500]
Y_rep <- posterior_predict_dbn(fit_static, ndraws = 500)
```

## 7  Wrap-up

* We simulated a single-time-point network,
* ran a short chain,
* checked convergence **first**,
* inspected parameters and latent effects, and
* used posterior-predictive ECDF / density overlays to evaluate fit.

Have some fun.