---
title:  "Dynamic bilinear network models"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Dynamic bilinear network models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE, comment = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = FALSE, warning = FALSE,
  eval      = NOT_CRAN
)
library(dbn)
library(ggplot2)
```

## 1 The model

The dynamic bilinear model allows the influence matrices $A_t$ and $B_t$
to change over time, capturing settings where the rules governing
network evolution are themselves evolving:

$$\Theta_t = A_t \, \Theta_{t-1} \, B_t^\top + M + \varepsilon_t$$

Entry $a_{i,i',t}$ of the sender influence matrix $A_t$ measures how
strongly actor $i'$'s past sending behavior predicts actor $i$'s current
sending at time $t$.  The receiver matrix $B_t$ captures the analogous
pattern on the targeting side.  The baseline mean $M$ absorbs stable
dyad-specific tendencies that persist regardless of the dynamics.

Two additional parameters govern how rapidly the influence structure
reorganizes.  The innovation variances $\tau_A^2$ and $\tau_B^2$ control
step-to-step variation in $A_t$ and $B_t$: values near zero indicate a
nearly static structure, while larger values allow rapid reorganization.
When `ar1 = TRUE`, the model adds persistence parameters $\rho_A$ and
$\rho_B$ that measure mean reversion.  Values of $\rho$ near 1 indicate
that the influence structure is highly persistent; values near 0 indicate
rapid reversion to the prior mean.

## 2 Simulate data

We generate a Gaussian network with time-varying influence matrices.
Setting `ar1 = TRUE` gives the influence matrices AR(1) dynamics, so each
period's influence structure is a noisy version of the previous period's
rather than an uncorrelated random walk.  The high persistence
($\rho = 0.9$) and small innovation variance ($\tau^2 = 0.04$) create
smoothly evolving dynamics that are easier to recover from limited data.

```{r simulate}
sim = simulate_dynamic_dbn(
  n      = 8,
  p      = 1,
  time   = 15,
  sigma2 = 0.3,
  tauA2  = 0.04,
  tauB2  = 0.04,
  ar1    = TRUE,
  rhoA   = 0.9,
  rhoB   = 0.9,
  seed   = 6886
)

Y = sim$Z
dim(Y)
```

## 3 Fit the model

We set `ar1 = TRUE` so the model uses AR(1) dynamics for $A_t$ and
$B_t$, and `update_rho = TRUE` so the persistence parameters are
estimated from the data rather than fixed at a default value.

```{r fit, cache = TRUE}
fit = dbn(
  Y,
  family     = "gaussian",
  model      = "dynamic",
  nscan      = 1000,
  burn       = 500,
  odens      = 2,
  ar1        = TRUE,
  update_rho = TRUE,
  verbose    = FALSE
)
```

## 4 Convergence diagnostics

The trace plots should show stationary chains with no drift.  For the
dynamic model, the key parameters are $\sigma^2$ (process noise),
$\tau_A^2$ and $\tau_B^2$ (innovation variances), and $\rho_A$ and
$\rho_B$ (persistence).

```{r convergence}
check_convergence(fit)
```

```{r trace, fig.height = 5}
plot_trace(fit, pars = c("sigma2", "tau_A2", "tau_B2", "rho_A", "rho_B"))
```

## 5 Parameter recovery

### Scalar parameters

```{r params}
param_summary(fit)
```

The model estimates variance parameters that govern the dynamics.  The
internal parameterization may differ in scale from the simulation inputs,
so direct comparison of scalar values is not always meaningful.  The most
informative validation is whether the model recovers the latent network
trajectories.

### Latent network recovery

The model's primary output is the latent network state $\Theta_t$.  We
compare the posterior mean of $\Theta_t + M$ to the true values across
all dyads and time points.  Strong correlation along the diagonal
indicates that the model is capturing both the cross-sectional structure
and the temporal dynamics.

```{r theta-recovery, fig.height = 4.5}
Theta_mean = apply(fit$Theta, 1:4, mean)
M_mean = apply(fit$M, 1:3, mean)

n_act = dim(sim$Theta)[1]
Tt = dim(sim$Theta)[4]

true_vals = numeric(0)
est_vals  = numeric(0)

for (t in seq_len(Tt)) {
  true_mat = sim$Theta[, , 1, t]
  est_mat  = Theta_mean[, , 1, t] + M_mean[, , 1]
  mask = !is.na(true_mat)
  true_vals = c(true_vals, true_mat[mask])
  est_vals  = c(est_vals, est_mat[mask])
}

df_theta = data.frame(truth = true_vals, estimate = est_vals)
r_sq = round(cor(df_theta$truth, df_theta$estimate)^2, 3)

ggplot(df_theta, aes(x = truth, y = estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.1, size = 0.8) +
  labs(
    title    = "Recovery of Latent Network State",
    subtitle = paste0("R-Squared = ", r_sq, " (All Dyads x Time Points)"),
    x = expression("True " * Theta[t]),
    y = expression("Posterior Mean " * Theta[t] + M)
  ) +
  theme_bw() +
  theme(panel.border = element_blank())
```

Each point represents one off-diagonal dyad at one time point.  Some
scatter is expected due to posterior shrinkage and estimation uncertainty,
but the bulk of points should cluster along the diagonal.

## 6 Dyad trajectories

The `dyad_path()` function tracks how a specific bilateral relationship
evolves over time, with a ribbon showing the posterior credible interval.
Wider ribbons indicate greater uncertainty about the trajectory of that
particular relationship.

```{r dyad-path, fig.height = 3.5}
dyad_path(fit, i = 1, j = 2)
```

## 7 Forecasting

We generate forecasts 5 steps ahead by propagating the estimated dynamics
forward.  Each of the 200 posterior draws produces a different forecast
trajectory (reflecting uncertainty in $A$, $B$, and $\sigma^2$), and
`summary = "mean"` averages across them.

```{r forecast}
fc = predict(fit, H = 5, S = 200, summary = "mean")
dim(fc)
```

The resulting array has dimensions `[actors, actors, relations, horizon]`.
These forecasts assume the final-period dynamics continue unchanged, so
they are most informative over short horizons.

## 8 Network statistics over time

The `network_summary()` function tracks a network-level statistic across
time with posterior credible intervals, providing a macro-level view of
how the network evolves.

```{r network-stats}
ns = network_summary(fit, stat = "density")
head(ns)
```

The `edge_prob()` function computes the posterior probability that each
edge is positive at a given time point.  Values near 1 indicate reliably
positive edges; values near 0.5 indicate substantial uncertainty about
the sign.

```{r edge-prob}
ep = edge_prob(fit, rel = 1, time = dim(Y)[4])
round(ep[1:5, 1:5], 2)
```

## 9 Bipartite networks

The package automatically detects bipartite networks when the sender and
receiver dimensions differ.  Passing a rectangular array causes the model
to estimate $A$ as `n_row x n_row` (sender dynamics) and $B$ as
`n_col x n_col` (receiver dynamics).

```{r bipartite, cache = TRUE}
sim_bp = simulate_dynamic_dbn(
  n = 8, n_col = 6, p = 1, time = 12, seed = 6886
)

fit_bp = dbn(
  sim_bp$Z,
  model   = "dynamic",
  family  = "gaussian",
  nscan   = 1000,
  burn    = 500,
  odens   = 2,
  verbose = FALSE
)
summary(fit_bp)
```

## 10 Parallel computation

The dynamic model parallelizes row-wise FFBS updates via OpenMP.  For
networks with 15 or more actors, multi-threading produces meaningful
speedups (typically 2-4x with 4-8 cores).

```{r threads, eval = FALSE}
set_dbn_threads(parallel::detectCores(logical = FALSE))
```

## 11 Scaling guidance

For larger networks, the main constraint is RAM.  Use `estimate_memory()`
before fitting, and adjust `time_thin` and `odens` to manage memory.

```{r memory-est}
estimate_memory(
  n_row = 50, n_col = 50, p = 1, Tt = 30,
  nscan = 1000, burn = 500, odens = 2,
  time_thin = 2, family = "gaussian"
)
```

For 100+ actors, expect to need 16-32 GB depending on settings.
Increasing `time_thin` and `odens` are the most effective ways to reduce
memory usage.

## 12 Next steps

For impulse response analysis that traces how shocks propagate through
the estimated dynamics, see `vignette("impulse_response")`.  For models
with known structural breaks where the influence structure shifts at
discrete time points, see `vignette("piecewise_dbn")`.  For large
networks (50+ actors), see `vignette("lowrank_dbn")`.
