---
title:  "Regime-switching (HMM) bilinear network models"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Regime-switching (HMM) bilinear network models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = FALSE, warning = FALSE,
  eval      = NOT_CRAN
)
library(dbn)
library(ggplot2)
```

## 1 Overview

The HMM variant of the dynamic bilinear network model allows for
regime-switching dynamics.  At each time point, the network is governed
by one of $R$ discrete regimes, each with its own influence matrices
$A_r$ and $B_r$.  Transitions between regimes follow a Markov chain with
an estimated transition matrix $\Pi$:

$$\Theta_t = A_{s_t} \, \Theta_{t-1} \, B_{s_t}^\top + M + \varepsilon_t, \quad s_t \sim \text{Markov}(\Pi)$$

This model is appropriate when network dynamics exhibit abrupt
structural breaks and the analyst wants the model to discover when those
breaks occur, rather than specifying them in advance.  For known break
points, the piecewise model provides more stable inference with fewer
parameters (see `vignette("piecewise_dbn")`).

Regime identification improves with network size, time series length, and
the magnitude of the structural difference between regimes.  For small
networks ($n < 15$) or subtle regime changes, the piecewise model with
known break points is a more reliable choice.

## 2 Simulate a regime-switching network

We simulate a two-regime network with 12 actors over 30 time periods.
The high `transition_prob` (0.9) means each regime persists for several
consecutive periods before switching.  The large `tau_A2` and `tau_B2`
values (0.8) create regimes with substantially different dynamics, which
is important for reliable regime identification.

```{r simulate, cache = TRUE}
sim = simulate_hmm_dbn(
  n     = 12,
  p     = 1,
  time  = 30,
  R     = 2,
  sigma2        = 0.3,
  tau_A2        = 0.8,
  tau_B2        = 0.8,
  transition_prob = 0.9,
  seed  = 6886
)

dim(sim$Y)

# true regime sequence and frequencies
sim$S
table(sim$S)
```

The true regime labels show which regime generated each time point.  With
two regimes and a transition probability of 0.9, we expect each regime
to persist for an average of 10 consecutive periods before switching.

## 3 Fit the HMM model

We set `model = "hmm"` and specify `R = 2` regimes.  The sampler jointly
estimates regime-specific dynamics matrices ($A_r$, $B_r$), the
transition matrix $\Pi$, and the regime labels $s_t$.

```{r fit, cache = TRUE}
fit = dbn(
  sim$Y,
  model   = "hmm",
  family  = "ordinal",
  R       = 2,
  nscan   = 1000,
  burn    = 500,
  odens   = 2,
  verbose = FALSE
)
```

## 4 Model summary and transition matrix

The summary reports the posterior mean transition matrix $\Pi$ and
variance parameters.  The diagonal of $\Pi$ indicates regime persistence:
values near 1 mean regimes tend to stick, while large off-diagonal values
indicate frequent switching.

```{r summary}
summary(fit)
```

Two things to note about the estimated $\Pi$.  First, the model's regime
labels may not match the simulation's — this is the standard
label-switching issue in mixture models and does not affect the
substantive results.  Second, with the moderate chain length used here
(`nscan = 1000`), the transition matrix has substantial posterior
uncertainty.  Recovering the true persistence (0.9 on the diagonal)
requires longer chains and larger networks; with $n = 12$ actors and
30 time points, the model has limited data per regime to pin down $\Pi$
precisely.  For substantive work, run at least `nscan = 5000` with
`burn = 2000`.

## 5 Regime probabilities

The posterior probability of each regime at each time point provides the
model's classification of the time series into regimes.  Probabilities
near 0 or 1 indicate confident classification; values near 0.5 indicate
ambiguity.  Regime identification depends on how different the regimes
are (controlled by `tau_A2` and `tau_B2` in the simulation) and how much
data is available per time point (number of actors).

```{r regime-probs, fig.height = 3.5}
plot_regime_probs(fit)
```

The `regime_probs()` function returns the underlying probability matrix
for custom analysis.  Each row sums to 1 across regimes.

```{r regime-table}
probs = regime_probs(fit)
head(probs, 10)
```

## 6 Model diagnostics

The default `plot()` method for HMM fits provides a combined diagnostic
view: regime probabilities, the estimated transition matrix, and MCMC
trace plots for monitoring convergence.

```{r plot-hmm, fig.height = 8}
plot(fit)
```

## 7 Dyad trajectory

The `dyad_path()` function tracks a specific bilateral relationship
over time with posterior credible intervals, averaging across regime
assignments.

```{r dyad, cache = TRUE, fig.height = 3.5}
dyad_path(fit, i = 2, j = 5)
```

## 8 Forecasting

The `predict()` method samples future regime states from $\Pi$ and
propagates the regime-specific bilinear dynamics forward.  This means
forecast uncertainty reflects both posterior uncertainty in $A_r$ and
$B_r$ and uncertainty about which regime will be active in future periods.

```{r forecast}
pred = predict(fit, H = 5, draws = 100, summary = "mean")
dim(pred)
```

## 9 When to use HMM

The HMM model is appropriate when network dynamics exhibit discrete
structural breaks (such as policy changes, crises, or regime
transitions) and the analyst wants data-driven discovery of when those
breaks occur.  The number of regimes $R$ should be small (2-5), and
the network should have enough actors and time points to identify
regime structure.  For smoothly evolving dynamics without discrete
breaks, `model = "dynamic"` is a better choice.  For known break points,
`model = "piecewise"` provides more stable inference.  For large networks
(50+ actors), `model = "lowrank"` offers better scalability.
