---
title:  "Impulse response analysis for network models"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Impulse response analysis for network models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = FALSE, warning = FALSE,
  eval      = NOT_CRAN
)
library(dbn)
library(ggplot2)
```

## 1 Overview

Impulse response functions (IRFs) measure how a shock to one part of the
network propagates through the bilinear dynamics over time.  Given the
model $\Theta_t = A_t \Theta_{t-1} B_t' + M + \varepsilon_t$, we inject
a perturbation $S$ at time $t_0$ and track a network-level statistic over
$H$ horizons:

$$\Delta_0 = S, \quad \Delta_h = A_{t_0+h} \, \Delta_{h-1} \, B_{t_0+h}'$$

At each horizon we compare a network statistic (density, degree, etc.)
computed on the perturbed versus baseline network.  The shock is applied
once; subsequent horizons show how the estimated dynamics propagate it
forward.

## 2 Fit a well-converged dynamic model

IRFs require a fitted model with time-varying $A$ and $B$ matrices.  We
use the Gaussian family for clean interpretation and simulate a 10-actor
network with small innovation variances to keep the dynamics stable.

```{r fit, cache = TRUE}
sim = simulate_dynamic_dbn(
  n = 10, p = 1, time = 15,
  sigma2 = 0.3, tauA2 = 0.03, tauB2 = 0.03,
  seed = 6886
)

fit = dbn(
  sim$Z,
  model   = "dynamic",
  family  = "gaussian",
  nscan   = 1000,
  burn    = 500,
  odens   = 2,
  verbose = FALSE
)
```

Always verify convergence before computing IRFs, since poorly converged
chains produce unreliable credible intervals:

```{r check-conv}
check_convergence(fit)
```

## 3 Building shock matrices

The `build_shock()` function creates structured perturbation matrices.
Each type addresses a different counterfactual question.

```{r shocks}
m = 10

# shock a single directed edge (actor 1 -> actor 2)
S_edge = build_shock(m, type = "unit_edge", i = 1, j = 2, magnitude = 2)

# shock all outgoing edges from actor 1
S_node = build_shock(m, type = "node_out", i = 1, magnitude = 1)

# uniform shock to all edges
S_dens = build_shock(m, type = "density", magnitude = 0.5)

# custom block shock (e.g., a group-level perturbation)
S_custom = matrix(0, m, m)
S_custom[1:3, 4:6] = 1.0
```

## 4 Edge shock: effect on network density

How does shocking a single edge affect overall network connectivity?
We add 2 units to the $(1,2)$ entry and track density (mean of
off-diagonal entries) over 5 horizons.

```{r irf-density, cache = TRUE, fig.height = 3.5}
irf_dens = compute_irf(
  fit,
  shock    = S_edge,
  H        = 5,
  t0       = 3,
  stat_fun = dbn::stat_density,
  n_draws  = 100
)

plot(irf_dens, title = "Edge Shock (1->2, +2) -> Network Density")
print(irf_dens)
```

At horizon 0, the effect is exact: adding 2 to one off-diagonal cell in a
10-actor network changes density by $2 / (10 \times 9) \approx 0.022$.  At
subsequent horizons, credible intervals widen as posterior uncertainty in
the transition matrices compounds.  The decay rate reflects the spectral
properties of the estimated $A_t$: eigenvalues near unity sustain the
perturbation, while values well below unity produce rapid decay.

## 5 Node shock: effect on out-degree

What happens when an actor increases all outgoing ties at once?  The
`node_out` shock sets all entries in row $i$ (including the self-loop
diagonal entry) to the specified magnitude.

```{r irf-degree, cache = TRUE, fig.height = 3.5}
irf_deg = compute_irf(
  fit,
  shock    = S_node,
  H        = 5,
  t0       = 3,
  stat_fun = function(X) stat_out_degree(X)[1],
  n_draws  = 100
)

plot(irf_deg, title = "Node-Out Shock (Actor 1, +1) -> Actor 1 Out-Degree")
```

Actor 1's out-degree jumps by 10 at horizon 0 (the shock sets all 10
entries in row 1, including the diagonal, to magnitude 1).  The posterior
mean of the effect decays toward zero at subsequent horizons, but the
credible intervals widen substantially as posterior uncertainty in the
transition matrices compounds across horizons.  Wide credible intervals
at longer horizons are typical and reflect how estimation uncertainty
accumulates in multi-step propagation.

## 6 Edge shock: effect on reciprocity

Reciprocity is a nonlinear statistic (the correlation between $X_{ij}$ and
$X_{ji}$).  Unlike density, it depends on the baseline network structure,
so there is posterior uncertainty even at horizon 0: the same shock
produces different reciprocity changes depending on the baseline network,
which varies across posterior draws.

```{r irf-recip, cache = TRUE, fig.height = 3.5}
irf_recip = compute_irf(
  fit,
  shock    = S_edge,
  H        = 5,
  t0       = 3,
  stat_fun = stat_reciprocity,
  n_draws  = 100
)

plot(irf_recip, title = "Edge Shock (1->2, +2) -> Network Reciprocity")
```

The nonzero credible interval at horizon 0 illustrates this distinction:
for linear statistics like density, the horizon-0 effect is a deterministic
function of the shock matrix, while for nonlinear statistics it interacts
with the posterior distribution of the baseline network.

## 7 Comparing multiple shocks

Combining IRF results in a single figure allows side-by-side comparison
across different shock types and statistics.  Each panel uses its own
y-axis scale because the statistics are on different scales.

```{r compare-irfs, fig.width = 7, fig.height = 4}
irfs = list(
  "Edge -> Density"     = irf_dens,
  "Node -> Out-Degree"  = irf_deg,
  "Edge -> Recipr."     = irf_recip
)

df_all = do.call(rbind, lapply(names(irfs), function(nm) {
  d = irfs[[nm]]
  data.frame(
    horizon = d$horizon,
    mean    = d$mean,
    q025    = d$q025,
    q975    = d$q975,
    shock   = nm
  )
}))

ggplot(df_all, aes(x = horizon, y = mean)) +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ shock, scales = "free_y") +
  labs(x = "Horizon", y = "IRF", title = "Impulse Response Comparison") +
  theme_bw() +
  theme(
    panel.border = element_blank(),
    strip.background = element_rect(fill = "black", color = "black"),
    strip.text.x = element_text(color = "white", hjust = 0)
  )
```

The key pattern to look for across all panels: a clear initial effect that
decays toward zero, with credible intervals that widen but do not explode.
An expanding credible interval that remains bounded is consistent with a
well-identified model; explosive intervals suggest the estimated dynamics
are unstable.

## 8 Interpreting IRFs

Horizon 0 captures the immediate, mechanical consequence of the shock.
For linear statistics (density, degree), there is no posterior uncertainty
at horizon 0 because the shock is a fixed constant.  For nonlinear
statistics (reciprocity, transitivity), there is uncertainty even at
horizon 0 because the statistic interacts with the baseline network, which
varies across posterior draws.

At horizons 1 and beyond, the estimated dynamics propagate the shock
forward.  Uncertainty widens because the transition matrices $A_t$ and
$B_t$ are themselves estimated with posterior noise.  The decay rate is
governed by the spectral radius of $A$: eigenvalues near unity sustain
perturbations, while values well below unity produce rapid decay.

It is important to note that IRFs are counterfactual scenarios that assume
the transition dynamics remain invariant to the shock.  The credible
intervals reflect posterior uncertainty in model parameters, not sampling
variability in the observed data.

## 9 Available network statistics

The `stat_fun` argument accepts any function that maps a matrix to a
scalar.  Built-in statistics include `stat_density()` (mean of
off-diagonal entries), `stat_out_degree()` and `stat_in_degree()` (row
and column sums), `stat_reciprocity()` (correlation between $X_{ij}$ and
$X_{ji}$), and `stat_transitivity()` (clustering coefficient).  For
degree-based statistics, wrap to extract a specific actor:
`function(X) stat_out_degree(X)[i]`.

## 10 Settings for publication

The examples above use moderate settings for vignette build times.  For
substantive analysis, increase both the model chain length and the number
of IRF draws to ensure stable credible intervals:

```{r pub-settings, eval = FALSE}
fit = dbn(Y, model = "dynamic", family = "gaussian",
          nscan = 10000, burn = 5000, odens = 5)

irf = compute_irf(fit, shock = S, H = 10, t0 = 5,
                  stat_fun = dbn::stat_density, n_draws = 500)
```

## 11 Next steps

For a complete applied workflow with real data and IRFs, see
`vignette("applied_ir")`.  For piecewise models that compare pre-
and post-regime influence structures, see `vignette("piecewise_dbn")`.
