---
title:  "Low-rank bilinear network models"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Low-rank bilinear network models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = FALSE, warning = FALSE,
  eval      = NOT_CRAN
)
library(dbn)
library(ggplot2)
```

## 1 Overview

The low-rank model factorizes the sender dynamics matrix as:

$$A_t = U \, \text{diag}(\alpha_t) \, U^\top$$

where $U$ is an $n \times r$ orthogonal matrix (constrained to the
Stiefel manifold) and $\alpha_t$ is a vector of time-varying factor
strengths.  This reduces the number of free parameters from $n^2$ to
$n \times r + r$.  For a 50-actor network, the full dynamic model
requires estimating 2,500 entries per time point, while a rank-3
low-rank model requires only 153 ($50 \times 3 + 3$).

The columns of $U$ define latent groups of actors that share similar
influence patterns.  The factor trajectories $\alpha_t$ capture how
strongly each latent dimension drives the dynamics at each time point.
This structure makes the model tractable for larger networks where the
full dynamic model becomes prohibitively expensive.

## 2 When to use low-rank

The low-rank model is designed for networks with 50 or more actors where
the full dynamic model is computationally prohibitive.  The parameter
savings grow quadratically with network size:

| Network Size | Full Dynamic Parameters | Rank-3 Low-Rank Parameters |
|:---:|:---:|:---:|
| 20 actors | 400 | 63 |
| 50 actors | 2,500 | 153 |
| 100 actors | 10,000 | 303 |

For smaller networks where full flexibility is affordable, use
`model = "dynamic"`.  For known structural break points, use
`model = "piecewise"`.

## 3 Memory estimation

Before fitting, check RAM requirements with `estimate_memory()`.  This
is particularly important for the low-rank model because its primary use
case is larger networks where memory is the binding constraint.

```{r memory-est}
estimate_memory(
  n_row = 50, n_col = 50, p = 1, Tt = 30,
  nscan = 5000, burn = 2000, odens = 5,
  family = "ordinal"
)
```

## 4 Simulate data

We simulate a 10-node network with rank $r = 2$, meaning the sender
dynamics are driven by two latent factors.  We use the Gaussian family
(`sim$Z`) for this demonstration because it provides sharper factor
identification than the ordinal family at small network sizes.  The
model's primary use case is networks with 50+ actors where the ordinal
family works well; this small example illustrates the API and factor
trajectory interpretation.

```{r simulate, cache = TRUE}
sim = simulate_lowrank_dbn(
  n          = 10,
  p          = 1,
  time       = 15,
  r          = 2,
  sigma2     = 0.3,
  tau_alpha2 = 0.1,
  tauB2      = 0.04,
  seed       = 6886
)

dim(sim$Z)
```

## 5 Fit the model

The key argument is `r`, the rank of the factorization.  Setting `r`
much smaller than `n` is what makes this model scalable.  In practice,
$r = 2$ or $r = 3$ captures the dominant dimensions of variation in the
influence structure for most applications.

```{r fit, cache = TRUE}
fit = dbn(
  sim$Z,
  model   = "lowrank",
  family  = "gaussian",
  r       = 2,
  nscan   = 1000,
  burn    = 500,
  odens   = 2,
  verbose = FALSE
)
```

## 6 Model diagnostics

The default `plot()` method provides a combined diagnostic view: MCMC
trace plots for scalar parameters, estimated factor trajectories
($\alpha_k$ over time), and the posterior mean of the node loading
matrix $U$.

```{r plot-model, fig.height = 8}
plot(fit)
```

## 7 Factor trajectories

The `tidy_dbn_lowrank()` function extracts factor trajectories in
tidy format for custom visualization.  Ribbons show 95% posterior
credible intervals.

```{r factor-traj, fig.height = 3.5}
alpha_df = tidy_dbn_lowrank(fit)

ggplot(alpha_df, aes(x = time, y = mean)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  facet_wrap(
    ~ factor, scales = "free_y",
    labeller = label_bquote(alpha[.(factor)])
  ) +
  labs(
    title = "Estimated Factor Trajectories",
    x = "Time", y = expression(alpha)
  ) +
  theme_bw() +
  theme(
    panel.border = element_blank(),
    strip.background = element_rect(fill = "black", color = "black"),
    strip.text.x = element_text(color = "white", hjust = 0)
  )
```

## 8 Model summary

```{r summary}
summary(fit)
```

## 9 Forecasting

Forecasts propagate both the factor dynamics $\alpha_t$ and the receiver
dynamics $B_t$ forward, generating forecast distributions that reflect
uncertainty in all model parameters.

```{r forecast}
pred = predict(fit, H = 3, draws = 100, summary = "mean")
dim(pred)
```

## 10 Choosing the rank

Start with $r = \lceil \log_2(n) \rceil$ and increase if posterior
predictive checks show poor fit.  Compare models with different ranks
using `compare_dbn()`.  Factors whose $\alpha_t$ trajectories remain
near zero throughout the time series contribute little and suggest a
lower rank suffices.

## 11 Next steps

For the full dynamic model without low-rank constraints, see
`vignette("dynamic_dbn")`.  For impulse response analysis, see
`vignette("impulse_response")`.  For the mathematical framework
underlying all model variants, see `vignette("methodology")`.
