---
title:  "Piecewise-static models for structural change"
author: "Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Piecewise-static models for structural change}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = FALSE, warning = FALSE,
  eval      = NOT_CRAN
)
library(dbn)
library(ggplot2)
```

## 1 Overview

Many political processes exhibit structural breaks, discrete moments when
the rules governing network dynamics fundamentally change.  The 2008
financial crisis reshuffled economic dependencies.  The Arab Spring
reorganized regional alliance patterns.  A leadership transition can
redirect a country's foreign policy overnight.

The piecewise-static model handles this setting: you specify when breaks
occurred and the model estimates how the influence structure differed
across regimes.  It fits block-constant influence matrices $A_k$ and
$B_k$ for each regime $k = 1, \ldots, K$:

$$\Theta_t = A_{k(t)} \, \Theta_{t-1} \, B_{k(t)}^\top + M + \varepsilon_t$$

The piecewise model sits between the static and dynamic extremes.  It
trades the time-point-level precision of the dynamic model for
interpretability (each regime has a single, directly comparable influence
matrix) and computational efficiency (estimating $K$ matrices instead of
$T-1$).

| Consideration | Piecewise | Dynamic |
|:---|:---:|:---:|
| Breaks are discrete and known | X | |
| Influence evolves smoothly | | X |
| Comparing pre/post periods | X | |
| Want interpretable regime summaries | X | |
| Memory constraints (large networks) | X | |
| Need time-point-specific estimates | | X |

## 2 Simulate data with a structural break

We simulate a network where the influence structure changes at a known
break point, then check whether the model recovers the difference.  The
`blocks` argument specifies the ending time index of each block: block 1
covers $t = 1$ to 15, block 2 covers $t = 16$ to 30.

```{r simulate, cache = TRUE}
sim = simulate_piecewise_dbn(
  n      = 8,
  time   = 30,
  blocks = c(15, 30),
  p      = 1,
  sigma2 = 0.5,
  tau2   = 0.3,
  seed   = 6886
)

dim(sim$Y)

# block structure
sim$block_info$K
sim$block_info$boundaries
sim$block_info$lengths

# average element-wise difference between regime A matrices
round(mean(abs(sim$true_A[[1]] - sim$true_A[[2]])), 3)
```

The element-wise difference between the two true $A$ matrices confirms
that the data-generating process produces genuinely different regimes.

## 3 Fit the piecewise model

Pass the known break points via `blocks`.  The model estimates separate
$A_k$ and $B_k$ for each regime while sharing the baseline mean $M$.

```{r fit, cache = TRUE}
fit = dbn(
  sim$Y,
  model   = "piecewise",
  family  = "ordinal",
  blocks  = c(15, 30),
  nscan   = 1000,
  burn    = 500,
  odens   = 2,
  verbose = FALSE
)

summary(fit)
```

## 4 Convergence diagnostics

```{r convergence, fig.height = 5}
check_convergence(fit)
plot_trace(fit, pars = c("s2", "t2", "g2"))
```

## 5 Regime comparison

The `compare_blocks()` function quantifies differences between regimes
with posterior uncertainty.  The output reports the posterior mean of the
Frobenius norm $\|A_k - A_{k+1}\|$, a 95% credible interval, and the
posterior probability that the difference exceeds a substantively
meaningful threshold (default 0.1).  A high probability indicates strong
evidence that the influence structure genuinely changed, not just
sampling noise.

```{r compare-blocks}
regime_diffs = compare_blocks(fit)
```

## 6 Extracting regime-specific influence

Each regime has its own estimated $A$ and $B$ matrices stored in the fit
object.

```{r regime-matrices}
A1_est = fit$A_blocks[[1]]
A2_est = fit$A_blocks[[2]]
```

The `compare_blocks()` result is the primary tool for assessing whether
the influence structure genuinely changed between regimes.  For
entry-level comparisons of individual $A$ elements, use a larger
network ($n \geq 20$).

## 7 Visualizing influence change

The difference $A_2 - A_1$ shows which sender-influence relationships
changed most across regimes.  Red entries indicate actors whose influence
increased in regime 2; blue entries indicate decreased influence.

```{r influence-change, fig.height = 5}
diff_A = A2_est - A1_est
n = nrow(diff_A)

df_heatmap = expand.grid(
  receiver = seq_len(n),
  sender   = seq_len(n)
)
df_heatmap$value = as.vector(diff_A)

ggplot(df_heatmap, aes(x = sender, y = receiver, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    name = expression(Delta * A)
  ) +
  labs(
    title = "Change in Sender Influence: Regime 2 vs. Regime 1",
    x = "Sender (Influenced By)", y = "Sender (Influences)"
  ) +
  coord_equal() +
  theme_bw() +
  theme(panel.border = element_blank())
```

## 8 Speed and memory advantages

The piecewise model estimates $K$ influence matrices instead of $T-1$,
pooling data within each regime.  This makes it faster and more
memory-efficient than the dynamic model, particularly for longer time
series.

```{r timing, cache = TRUE}
sim_t = simulate_piecewise_dbn(n = 10, time = 30, blocks = 3, seed = 6886)

t_pw = system.time({
  fit_pw = dbn(sim_t$Y, model = "piecewise", blocks = 3,
               nscan = 1000, burn = 500, verbose = FALSE)
})

t_dyn = system.time({
  fit_dyn = dbn(sim_t$Y, model = "dynamic",
                nscan = 1000, burn = 500, verbose = FALSE)
})

data.frame(
  model   = c("Piecewise", "Dynamic"),
  seconds = round(c(t_pw["elapsed"], t_dyn["elapsed"]), 1)
)
```

For large networks, use `store_theta = FALSE` to avoid storing the full
$\Theta$ posterior (which scales as $n^2 \times T \times \text{draws}$).
This retains $A$, $B$, $M$, variance posteriors, convergence
diagnostics, and `compare_blocks()` support while substantially reducing
memory.

```{r memory-compare}
fit_full = dbn(sim$Y, model = "piecewise", blocks = c(15, 30),
               nscan = 200, burn = 100, verbose = FALSE, store_theta = TRUE)
fit_lean = dbn(sim$Y, model = "piecewise", blocks = c(15, 30),
               nscan = 200, burn = 100, verbose = FALSE, store_theta = FALSE)

data.frame(
  storage = c("With Theta", "Without Theta"),
  mb      = round(c(object.size(fit_full), object.size(fit_lean)) / 1e6, 2)
)
```

## 9 Specifying blocks

The `blocks` argument is flexible.  A single integer creates that many
equal-sized blocks.  A vector of integers specifies the ending time index
of each block.  Named vectors improve readability for applied work.

```{r block-specs, eval = FALSE}
# integer: equal-sized regimes
fit = dbn(Y, model = "piecewise", blocks = 4)

# vector of endpoints
fit = dbn(Y, model = "piecewise", blocks = c(10, 25, 40))

# named vector for interpretability
fit = dbn(Y, model = "piecewise",
          blocks = c(pre_war = 15, war = 30, post_war = 45))
```

## 10 Gaussian family

The piecewise model supports all three outcome families.  For continuous
data, switch to `family = "gaussian"`:

```{r gaussian, cache = TRUE}
fit_gauss = dbn(
  sim$Y_continuous,
  model   = "piecewise",
  family  = "gaussian",
  blocks  = c(15, 30),
  nscan   = 1000,
  burn    = 500,
  odens   = 2,
  verbose = FALSE
)

summary(fit_gauss)
```

## 11 Applied example: UNGA voting and the 2008 financial crisis

This example examines UN General Assembly voting alignment using the
financial crisis as a structural break.  It requires the `peacesciencer`
package.  Because it uses external data, we show the code and
pre-computed output rather than evaluating inline.

```r
library(peacesciencer)

# build dyadic panel 1995-2015
dyads = create_dyadyears(subset_years = 1995:2015) |>
  add_fpsim()

# select ~100 countries with complete data
years = 1995:2015
complete_countries = Reduce(intersect, lapply(years, function(y) {
  unique(c(dyads$ccode1[dyads$year == y], dyads$ccode2[dyads$year == y]))
}))

country_coverage = table(c(
  dyads$ccode1[dyads$ccode1 %in% complete_countries],
  dyads$ccode2[dyads$ccode2 %in% complete_countries]
))
top_countries = as.numeric(
  names(sort(country_coverage, decreasing = TRUE))[1:100]
)

# build [n, n, 1, T] array of kappavv similarity scores
dyads_sub = dyads[dyads$ccode1 %in% top_countries &
                    dyads$ccode2 %in% top_countries, ]
actor_codes = sort(unique(c(dyads_sub$ccode1, dyads_sub$ccode2)))
n_actors = length(actor_codes)
code_to_idx = setNames(seq_along(actor_codes), actor_codes)

Y_unga = array(NA_real_, dim = c(n_actors, n_actors, 1, length(years)))
for (i in seq_len(nrow(dyads_sub))) {
  row_i = code_to_idx[as.character(dyads_sub$ccode1[i])]
  col_j = code_to_idx[as.character(dyads_sub$ccode2[i])]
  t_idx = which(years == dyads_sub$year[i])
  Y_unga[row_i, col_j, 1, t_idx] = dyads_sub$kappavv[i]
}
for (t in seq_along(years)) diag(Y_unga[, , 1, t]) = NA

# fit piecewise model with 2008 crisis as break point
t_break = which(years == 2008)
fit_crisis = dbn(
  Y_unga,
  model  = "piecewise",
  family = "gaussian",
  blocks = c(t_break, length(years)),
  nscan  = 5000,
  burn   = 2000,
  odens  = 5,
  store_theta = FALSE,
  verbose = TRUE
)

# did the influence structure change?
compare_blocks(fit_crisis)
#> Block Comparison Results (A)
#> block_1 vs block_2: ||dA|| = 2.34 [1.89, 2.91]
#>   P(||dA|| > 0.1) = 1

# which actors' influence changed most?
A_pre  = fit_crisis$A_blocks[[1]]
A_post = fit_crisis$A_blocks[[2]]
influence_change = rowSums(abs(A_post)) - rowSums(abs(A_pre))
```

A high probability that the Frobenius norm exceeds the threshold
indicates the crisis genuinely altered alignment dynamics.  In typical
applications, emerging market countries from Latin America, Asia, and
Africa show increased influence on alignment patterns post-2008, while
G7 nations show flat or declining dynamic influence.  Countries that
actively built coalitions (Brazil, Turkey, South Africa) tend to emerge
as gainers.  Substantive conclusions require longer MCMC runs, robustness
checks with alternative break points, and domain expertise.

## 12 Scaling to large networks

| Actors | Theta Storage | Expected Runtime | Memory |
|:---:|:---:|:---:|:---:|
| 50 | OK | 1-5 min | < 1 GB |
| 100 | Marginal | 5-20 min | 2-8 GB |
| 200 | Disable | 30-90 min | 1-4 GB |
| 500+ | Disable | Hours | 5-20 GB |

For 200+ actors, always use `store_theta = FALSE`.  You retain $A$, $B$,
$M$, variance posteriors, convergence diagnostics, and
`compare_blocks()` support; you lose full $\Theta$ posterior uncertainty.

```{r large-settings, eval = FALSE}
fit_large = dbn(
  Y_large,
  model       = "piecewise",
  family      = "gaussian",
  blocks      = c(25, 50),
  store_theta = FALSE,
  nscan       = 5000,
  burn        = 2000,
  odens       = 5,
  verbose     = TRUE
)
```

## 13 Practical guidance

### Choosing break points

Good candidates include known events (war onset, treaty signing), policy
interventions (sanctions, trade agreements), and external shocks
(financial crises, pandemics).  If uncertain about break point location,
fit multiple models with different break specifications and compare with
`compare_dbn()`.

### How many regimes?

| Time Points | Max Regimes |
|:---:|:---:|
| 20-30 | 2-3 |
| 30-50 | 3-4 |
| 50-100 | 4-6 |

### Relationship to other models

Piecewise with $K = 1$ is equivalent to the static model.  Compared to
the dynamic model, piecewise trades time-point precision for
interpretability and speed.  Compared to the HMM model, piecewise
requires the analyst to specify break points rather than discovering them
from the data.

## 14 Next steps

For smoothly evolving dynamics, see `vignette("dynamic_dbn")`.  For
data-driven regime discovery, see `vignette("hmm_dbn")`.  For impulse
response analysis, see `vignette("impulse_response")`.  For a complete
applied workflow with IRFs, see `vignette("applied_ir")`.
