---
title:  "Getting started: static bilinear network model"
author: "Tosin Salau & Shahryar Minhas"
date:   "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started: static bilinear network model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 6, fig.height = 4,
  message   = FALSE, warning = FALSE,
  eval      = NOT_CRAN
)
library(dbn)
library(ggplot2)
```

## 1 Overview

Dynamic bilinear network (DBN) models estimate how past network
interactions predict future interactions.  The package takes as input a
time series of networks stored as a 4D array with dimensions
`[senders, receivers, relation_types, time_points]`, where diagonal
entries (self-ties) are set to `NA`.

The static model is the simplest variant.  It estimates time-invariant
sender and receiver influence matrices $A$ and $B$ that govern how the
latent network state evolves:

$$\Theta_t = A \, \Theta_{t-1} \, B^\top + M + \varepsilon_t$$

The parameter $A$ captures sender influence: entry $a_{i,i'}$ measures
how strongly actor $i'$'s past sending behavior predicts actor $i$'s
current sending.  $B$ captures the analogous pattern on the receiving
side.  $M$ absorbs persistent dyad-specific tendencies such as stable
rivalries or alliances.  Because $A$ and $B$ are constant across time,
the rules governing network evolution do not change; only the network
state $\Theta_t$ itself moves.

## 2 Simulate data

We simulate a small network so we have known ground truth to compare
against.  The simulation generates both ordinal data (`sim$Y`, integers
1 through 5) and continuous latent values (`sim$Z`).  We use `sim$Z`
here because the Gaussian family expects continuous input.

```{r simulate}
sim = simulate_static_dbn(
  n      = 12,   # 12 actors
  p      = 1,    # 1 relation type
  time   = 15,   # 15 time periods
  sigma2 = 0.5,  # process noise variance
  tau2   = 0.1,  # prior variance for A and B
  seed   = 6886
)

Y = sim$Z
dim(Y)
```

The output confirms the expected `[12, 12, 1, 15]` array structure.  The
simulation also returns the true $A$, $B$, $M$, and $\Theta$ matrices,
which we use below to verify that the model recovers the data-generating
process.

## 3 Fit the model

We pass the data array to `dbn()` and specify the model type and outcome
family.  Three MCMC settings control sampling: `nscan` sets the number of
posterior samples to draw after burn-in, `burn` sets the warm-up period
to discard, and `odens` thins the chain by saving every $k$-th sample
to reduce autocorrelation.

```{r fit, cache = TRUE}
fit = dbn(
  Y,
  model   = "static",
  family  = "gaussian",
  nscan   = 1000,
  burn    = 500,
  odens   = 2,
  verbose = FALSE
)
```

## 4 Convergence diagnostics

MCMC methods produce samples from the posterior distribution, but those
samples are only useful if the chain has converged, meaning it has found
and is exploring the right region of parameter space.  Well-behaved
trace plots show stationary fluctuation around a stable mean.  Drift,
the chain getting stuck, or slow oscillation are all signs that the
chain needs to run longer.

```{r convergence, fig.height = 5}
check_convergence(fit)
plot_trace(fit)
```

The diagnostics report effective sample size (ESS) and Geweke
statistics.  ESS values of at least a few hundred indicate adequate
mixing.  The three parameters shown are $\sigma^2$ (process variance),
$\tau^2$ (A/B prior variance), and $g^2$ (latent variance).  If any
trace shows trending or poor mixing, increase both `burn` and `nscan`
before proceeding.

## 5 Parameter recovery

A key advantage of simulation studies is that we can check whether the
model recovers the known truth.  The scalar variance parameters can
differ in scale between the simulation and the model's internal
parameterization, so the most informative check is whether the model
recovers the latent baseline structure $M$.

### Scalar parameters

```{r param-recovery}
param_summary(fit)
```

### Baseline mean M

The posterior mean of $M$ should correlate with the true $M$ from the
simulation.  We exclude diagonal entries (self-ties are `NA` by
construction) and compare all off-diagonal elements.

```{r m-recovery, fig.height = 4.5}
M_est = latent_summary(fit, fun = mean)

# exclude diagonal entries from both truth and estimate
n_act = 12
off_diag = which(M_est$i != M_est$j)
est_M_vec = M_est$value[off_diag]

true_M_mat = sim$M[, , 1]
diag(true_M_mat) = NA
true_M_vec = as.vector(true_M_mat)
true_M_vec = true_M_vec[!is.na(true_M_vec)]

df_M = data.frame(truth = true_M_vec, estimate = est_M_vec)
r_sq = round(cor(df_M$truth, df_M$estimate)^2, 3)

ggplot(df_M, aes(x = truth, y = estimate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.5) +
  labs(
    title = "Recovery of Baseline Mean M",
    subtitle = paste0("R-Squared = ", r_sq),
    x = "True M", y = "Posterior Mean M"
  ) +
  theme_bw() +
  theme(panel.border = element_blank())
```

Points clustering along the 45-degree line indicate that the model
is recovering the true baseline relational tendencies.  An $R^2$ above
0.8 suggests the posterior mean of $M$ captures most of the
cross-sectional structure in the data-generating process.

## 6 Model summary

The `summary()` method reports posterior means and credible intervals
for the scalar parameters, along with data dimensions, MCMC settings,
and the outcome family.

```{r summary}
summary(fit)
```

## 7 Posterior predictive check

A posterior predictive check asks whether data simulated from the fitted
model look like the observed data.  We generate replicated datasets from
the posterior, then compare their empirical CDF to the observed data.
Replicated curves that closely track the observed curve indicate the model
captures the data's distributional features.

```{r ppc, fig.height = 4}
ppd = posterior_predict_dbn(fit, ndraws = 20, seed = 6886)
plot_ppc_ecdf(fit, ppd, Y_obs = Y, rel = 1)
```

## 8 Other outcome families

The static model supports two additional families beyond Gaussian.
The ordinal family uses a rank likelihood that respects the ordering of
categorical data (event counts, Likert items) without assuming a
specific scale.  The binary family uses a probit link for 0/1 tie data.

```{r other-families, cache = TRUE}
# ordinal: for ranked data (e.g., event counts, Likert items)
fit_ord = dbn(sim$Y, model = "static", family = "ordinal",
              nscan = 1000, burn = 500, odens = 2, verbose = FALSE)
summary(fit_ord)

# binary: for 0/1 tie presence/absence
Y_binary = array(as.integer(sim$Y[, , , ] > 3), dim = dim(sim$Y))
for (t in seq_len(dim(Y_binary)[4])) diag(Y_binary[, , 1, t]) = NA
fit_bin = dbn(Y_binary, model = "static", family = "binary",
              nscan = 1000, burn = 500, odens = 2, verbose = FALSE)
summary(fit_bin)
```

## 9 Next steps

The static model assumes the influence structure is constant.  For
applications where those rules may change, the dynamic model allows
$A_t$ and $B_t$ to evolve over time (see `vignette("dynamic_dbn")`).
When discrete structural breaks are known in advance, the piecewise
model estimates separate influence matrices for each regime (see
`vignette("piecewise_dbn")`).  For the full mathematical framework,
see `vignette("methodology")`.
